---
title: "Third Places and Civic Life in Texas"
author: "Annette Strauss Institute for Civic Life, University of Texas at Austin"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: hide
    fig_width: 11
    fig_height: 7
    df_print: paged
---

<style>
.main-container {
  max-width: 1400px;
}
body {
  font-size: 11pt;
}
h1 {
  color: #BF5700;
  border-bottom: 2px solid #BF5700;
  padding-bottom: 10px;
}
h2 {
  color: #005F86;
}
.caption {
  font-style: italic;
  color: #666;
  font-size: 0.9em;
}
</style>

```{r setup, include=TRUE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.align = "center",
  cache = FALSE
)

# Load packages
library(tidyverse)
library(haven)
library(labelled)
library(survey)
library(scales)
library(knitr)
library(kableExtra)
library(patchwork)
library(forcats)
library(sf)
library(tigris)


# Set UT Austin branding colors
ut_orange <- "#BF5700"
ut_blue <- "#005F86"
ut_teal <- "#00A9B7"
ut_gray <- "#666666"

# Custom ggplot theme
theme_asi <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 13, color = ut_blue),
      plot.subtitle = element_text(size = 10, color = ut_gray, margin = margin(b = 10)),
      plot.caption = element_text(size = 8, color = ut_gray, hjust = 0),
      axis.title = element_text(size = 10, face = "bold"),
      legend.position = "bottom",
      legend.title = element_text(face = "bold", size = 9),
      strip.text = element_text(face = "bold", size = 10, color = ut_blue),
      strip.background = element_rect(fill = "gray95", color = NA),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "gray80", fill = NA)
    )
}

theme_set(theme_asi())

# Color palette for participation
part_colors <- c(
  "Voted" = "#2ecc71",
  "Registered, no vote" = "#f39c12", 
  "Not registered" = "#e74c3c"
)

# Helper: Create publication-ready tables with sample sizes
nice_table <- function(df, title = NULL, digits = 2, include_n = TRUE) {
  raw_names <- names(df)
  pretty_names <- raw_names %>%
    str_replace_all("_", " ") %>%
    str_replace("^pct ", "") %>%
    str_replace("^pct", "") %>%
    str_replace("^n ", "N ")
  pct_cols <- str_detect(raw_names, "^pct[_ ]")
  pretty_names[pct_cols] <- paste0(pretty_names[pct_cols], " (%)")
  pretty_names[str_detect(raw_names, "^n[_ ]")] <- str_replace(pretty_names[str_detect(raw_names, "^n[_ ]")], "^n ", "N ")

  df %>%
    mutate(across(where(is.numeric), ~ round(., digits = digits))) %>%
    kable(
      caption = title, 
      col.names = pretty_names,
      format = "html",
      booktabs = TRUE,
      align = "c"
    ) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE,
      position = "left",
      font_size = 11
    ) %>%
    row_spec(0, bold = TRUE, color = "white", background = ut_blue)
}

# Helper: Format percentages
fmt_pct <- function(x, digits = 1) {
  paste0(round(x * 100, digits), "%")
}

```

```{r mapping-tx, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}

# Data 
tpi_path <- "data/tpi_ces_by_county1.csv"
tpi_total_path <- "data/third_place_index_1.0.csv"
tpi_cesdata <- read.csv(tpi_path)
tpi_total <- read.csv(tpi_total_path) 

# Create a Texas subset
tpi_tx <- tpi_cesdata %>%
  filter(state_fips == 48)

```

# Maps

## United States: Third Place Density by County

```{r us-map, echo=TRUE, message=FALSE, warning=FALSE, fig.width=12, fig.height=7}

# Get US county geometries
us_counties <- tigris::counties(cb = TRUE, year = 2021, progress_bar = FALSE) %>%
  st_transform(crs = 5070)  # Albers Equal Area projection for US

# Prepare TPI data with proper FIPS formatting
tpi_map_data <- tpi_cesdata %>%
  mutate(GEOID = str_pad(as.character(county_fips), 5, pad = "0")) %>%
  select(GEOID, total_places_per_sqmi, total_places, state_fips)

# Join spatial data with TPI data
us_counties_tpi <- us_counties %>%
  left_join(tpi_map_data, by = "GEOID")

# Create US map
ggplot(us_counties_tpi) +
  geom_sf(aes(fill = total_places_per_sqmi), color = NA, size = 0) +
  scale_fill_viridis_c(
    option = "plasma",
    name = "Third Places\nper Square Mile",
    trans = "log10",
    breaks = c(1, 10, 100, 1000),
    labels = c("1", "10", "100", "1,000"),
    na.value = "gray90"
  ) +
  labs(
    title = "Third Place Index: Density of Community Spaces Across US Counties",
    subtitle = "Total third places per square mile (log scale)",
    caption = "Source: OpenStreetMap data\nGray areas indicate missing data"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, color = ut_blue),
    plot.subtitle = element_text(size = 11, color = ut_gray),
    plot.caption = element_text(size = 9, color = ut_gray, hjust = 0),
    legend.position = "right",
    legend.title = element_text(face = "bold", size = 10),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

```

## Texas: Third Place Density by County

```{r texas-map, echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

# Get Texas county geometries
tx_counties <- tigris::counties(state = "TX", cb = TRUE, year = 2021, progress_bar = FALSE) %>%
  st_transform(crs = 3083)  # Texas Centric Albers Equal Area projection

# Prepare Texas TPI data with proper FIPS formatting
tpi_tx_map <- tpi_tx %>%
  mutate(GEOID = str_pad(as.character(county_fips), 5, pad = "0")) %>%
  select(GEOID, total_places_per_sqmi, total_places, state_fips)

# Join spatial data with TPI data for Texas
tx_counties_tpi <- tx_counties %>%
  left_join(tpi_tx_map, by = "GEOID")

# Create Texas map
ggplot(tx_counties_tpi) +
  geom_sf(aes(fill = total_places_per_sqmi), color = "white", size = 0.1) +
  scale_fill_viridis_c(
    option = "plasma",
    name = "Third Places\nper Square Mile",
    trans = "log10",
    breaks = c(1, 10, 100, 1000),
    labels = c("1", "10", "100", "1,000"),
    na.value = "gray90"
  ) +
  labs(
    title = "Third Place Index: Density of Community Spaces in Texas Counties",
    subtitle = "Total third places per square mile (log scale)",
    caption = "Source: OpenStreetMap data\nGray areas indicate missing data"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, color = ut_blue),
    plot.subtitle = element_text(size = 11, color = ut_gray),
    plot.caption = element_text(size = 9, color = ut_gray, hjust = 0),
    legend.position = "right",
    legend.title = element_text(face = "bold", size = 10),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

```

# Census Tract-Level Analysis

The following maps show third place density at the census tract level, providing much more granular detail than county-level aggregation. This reveals neighborhood-level patterns in community space availability.

## United States: Third Place Density by Census Tract

```{r us-tract-map, echo=TRUE, message=FALSE, warning=FALSE, fig.width=12, fig.height=7}

# Get US census tract geometries (this may take a moment)
us_tracts <- tigris::tracts(cb = TRUE, year = 2021, progress_bar = FALSE) %>%
  st_transform(crs = 5070)  # Albers Equal Area projection

# Prepare tract-level TPI data
tpi_tract_data <- tpi_total %>%
  mutate(GEOID = as.character(geoid)) %>%
  select(GEOID, total_places_per_sqmi, total_places, state_fips, third_place_percentile)

# Join spatial data with TPI data
us_tracts_tpi <- us_tracts %>%
  left_join(tpi_tract_data, by = "GEOID")

# Create US census tract map
ggplot(us_tracts_tpi) +
  geom_sf(aes(fill = third_place_percentile), color = NA, size = 0) +
  scale_fill_viridis_c(
    option = "plasma",
    name = "Third Place\nPercentile",
    limits = c(0, 1),
    breaks = seq(0, 1, 0.1),
    labels = c("0", "10", "20", "30", "40", "50", "60", "70", "80", "90", "100"),
    na.value = "gray90",
    guide = guide_colorbar(
      barwidth = 1.5,
      barheight = 15,
      ticks.colour = "black",
      ticks.linewidth = 0.5,
      frame.colour = "black"
    )
  ) +
  labs(
    title = "Third Place Index: Census Tract Percentile Rankings Across the United States",
    subtitle = "Third place percentile by census tract (0 = lowest density, 100 = highest density)",
    caption = "Source: OpenStreetMap data, ACS 2021\nGray areas indicate missing data. Each polygon represents a census tract."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, color = ut_blue),
    plot.subtitle = element_text(size = 11, color = ut_gray),
    plot.caption = element_text(size = 9, color = ut_gray, hjust = 0),
    legend.position = "right",
    legend.title = element_text(face = "bold", size = 10),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) 

```

## Texas: Third Place Density by Census Tract

```{r texas-tract-map, echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

# Get Texas census tract geometries
tx_tracts <- tigris::tracts(state = "TX", cb = TRUE, year = 2021, progress_bar = FALSE) %>%
  st_transform(crs = 3083)  # Texas-specific projection

# Prepare Texas tract-level TPI data
tpi_tx_tract <- tpi_total %>%
  filter(state_fips == 48) %>%
  mutate(GEOID = as.character(geoid)) %>%
  select(GEOID, total_places_per_sqmi, total_places, third_place_percentile, 
         urban_rural_class, total_population)

# Join spatial data with TPI data for Texas
tx_tracts_tpi <- tx_tracts %>%
  left_join(tpi_tx_tract, by = "GEOID")

# Create Texas census tract map
ggplot(tx_tracts_tpi) +
  geom_sf(aes(fill = third_place_percentile), color = NA, size = 0) +
  scale_fill_viridis_c(
    option = "plasma",
    name = "Third Place\nPercentile",
    limits = c(0, 1),
    breaks = seq(0, 1, 0.1),
    labels = c("0", "10", "20", "30", "40", "50", "60", "70", "80", "90", "100"),
    na.value = "gray90",
    guide = guide_colorbar(
      barwidth = 1.5,
      barheight = 15,
      ticks.colour = "black",
      ticks.linewidth = 0.5,
      frame.colour = "black"
    )
  ) +
  labs(
    title = "Third Place Index: Census Tract Percentile Rankings in Texas",
    subtitle = "Third place percentile by census tract (0 = lowest density, 100 = highest density)",
    caption = "Source: OpenStreetMap data, ACS 2021\nGray areas indicate missing data. 6,896 census tracts shown."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, color = ut_blue),
    plot.subtitle = element_text(size = 11, color = ut_gray),
    plot.caption = element_text(size = 9, color = ut_gray, hjust = 0),
    legend.position = "right",
    legend.title = element_text(face = "bold", size = 10),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

```

### Key Differences: Census Tract vs County Maps

**Granularity**: Census tracts reveal neighborhood-level variation that county maps obscure. For example, you can now see:

- Dense urban cores vs. suburban sprawl within the same county
- Variation between different neighborhoods in major cities
- Rural tract-by-tract differences in community space access

**Data Volume**: The US has ~83,500 census tracts vs. ~3,200 counties, providing 26× more spatial detail.

**Use Cases**: 

- Tract-level data is ideal for understanding intra-city inequality
- County-level data is better for regional comparisons and statewide policy analysis
