---
title: "Voter Participation in the 2024 U.S. General Election"
subtitle: "Analysis of CMPS 2024 Data with Focus on Texas"
author: "Annette Strauss Institute for Civic Life, University of Texas at Austin"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

<style>
.main-container {
  max-width: 1200px;
  margin-left: auto;
  margin-right: auto;
}
body {
  font-family: "Helvetica Neue", Helvetica, Arial, sans-serif;
  font-size: 14px;
  line-height: 1.6;
}
h1 {
  color: #BF5700;
  font-weight: bold;
}
h2 {
  color: #005F86;
  border-bottom: 2px solid #BF5700;
  padding-bottom: 5px;
}
.caption {
  font-size: 12px;
  font-style: italic;
  color: #666666;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load packages
library(tidyverse)
library(haven)
library(labelled)
library(survey)
library(scales)
library(knitr)
library(kableExtra)
library(patchwork)
library(forcats)

# Set UT Austin branding colors
ut_orange <- "#BF5700"
ut_blue <- "#005F86"
ut_teal <- "#00A9B7"
ut_gray <- "#666666"

# Custom ggplot theme
theme_asi <- function() {
  theme_minimal() +
    theme(
      text = element_text(family = "sans"),
      plot.title = element_text(size = 16, face = "bold", color = ut_orange),
      plot.subtitle = element_text(size = 12, color = ut_gray),
      axis.title = element_text(size = 11, face = "bold"),
      axis.text = element_text(size = 10),
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank()
    )
}

theme_set(theme_asi())

# Color palette for participation
part_colors <- c(
  "Voted" = ut_blue,
  "Did Not Vote" = ut_orange,
  "Democrats" = "#0015BC",
  "Republicans" = "#E81B23",
  "Independents" = "#9370DB"
)

# Helper: Create publication-ready tables with sample sizes
nice_table <- function(df, title = NULL, digits = 2, include_n = TRUE) {
  tbl <- df %>%
    kable(
      format = "html",
      digits = digits,
      caption = title,
      align = c("l", rep("r", ncol(df) - 1))
    ) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE,
      position = "left"
    ) %>%
    row_spec(0, bold = TRUE, color = "white", background = ut_orange)
  
  return(tbl)
}

# Helper: Format percentages
fmt_pct <- function(x, digits = 1) {
  paste0(format(round(x, digits), nsmall = digits), "%")
}
```

```{r load-data, cache=TRUE}
# Load CMPS 2024 data
cmps_path <- "cmps_data/cmps2024_fulladult_112425.csv"
cmps_raw <- read.csv(cmps_path)

# Data preparation
cmps_24 <- cmps_raw %>%
  mutate(
    # State and Texas indicator
    state = s21,
    texas = (state == 48),
    
    # Voting behavior (Q12)
    vote24 = case_when(
      q12 == 1 ~ "Did not vote",
      q12 == 2 ~ "Don't think I voted",
      q12 == 3 ~ "Think I voted",
      q12 == 4 ~ "Definitely voted",
      TRUE ~ NA_character_
    ),
    voted = case_when(
      q12 %in% c(3, 4) ~ 1,  # Think/definitely voted
      q12 %in% c(1, 2) ~ 0,  # Did not/don't think voted
      TRUE ~ NA_real_
    ),
    
    # Registration status (S27)
    registered = case_when(
      s27 == 1 ~ "Registered",
      s27 == 2 ~ "Not registered",
      s27 == 3 ~ "Not eligible",
      TRUE ~ NA_character_
    ),
    
    # Demographics
    # Age (using age5 for 5 categories)
    age_group = case_when(
      age5 == 1 ~ "18-29",
      age5 == 2 ~ "30-44",
      age5 == 3 ~ "45-59",
      age5 == 4 ~ "60-69",
      age5 == 5 ~ "70+",
      TRUE ~ NA_character_
    ),
    
    # Race/Ethnicity (qraceid)
    race_ethnicity = case_when(
      qraceid == 1 ~ "White",
      qraceid == 2 ~ "Hispanic/Latino",
      qraceid == 3 ~ "Black",
      qraceid == 4 ~ "Asian",
      qraceid == 5 ~ "Native American",
      qraceid == 6 ~ "MENA",
      qraceid == 7 ~ "NHPI",
      TRUE ~ NA_character_
    ),
    
    # Gender (S14)
    gender = case_when(
      s14 == 1 ~ "Male",
      s14 == 2 ~ "Female",
      s14 == 3 ~ "Non-binary",
      TRUE ~ NA_character_
    ),
    
    # Education (S17)
    education = case_when(
      s17 == 1 ~ "Grade 1-8",
      s17 == 2 ~ "Some HS",
      s17 == 3 ~ "HS diploma/GED",
      s17 == 4 ~ "Some college",
      s17 == 5 ~ "2-year degree",
      s17 == 6 ~ "4-year degree",
      s17 == 7 ~ "Postgraduate",
      TRUE ~ NA_character_
    ),
    education_cat = case_when(
      s17 %in% c(1, 2, 3) ~ "HS or less",
      s17 %in% c(4, 5) ~ "Some college",
      s17 %in% c(6, 7) ~ "4-year degree+",
      TRUE ~ NA_character_
    ),
    
    # Income (hhinc)
    income = case_when(
      hhinc == 1 ~ "Under $40k",
      hhinc == 2 ~ "$40k-$59k",
      hhinc == 3 ~ "$60k-$99k",
      hhinc == 4 ~ "$100k+",
      hhinc == 5 ~ "Prefer not to say",
      TRUE ~ NA_character_
    ),
    
    # Party ID (S28)
    party = case_when(
      s28 == 1 ~ "Democrat",
      s28 == 2 ~ "Republican",
      s28 == 3 ~ "Independent",
      s28 == 4 ~ "Other",
      TRUE ~ NA_character_
    ),
    
    # Ideology (Q8)
    ideology = case_when(
      q8 == 1 ~ "Very liberal",
      q8 == 2 ~ "Liberal",
      q8 == 3 ~ "Moderate (lean liberal)",
      q8 == 4 ~ "Moderate",
      q8 == 5 ~ "Moderate (lean conservative)",
      q8 == 6 ~ "Conservative",
      q8 == 7 ~ "Very conservative",
      q8 == 8 ~ "Don't think in these terms",
      TRUE ~ NA_character_
    ),
    
    # Feeling thermometers (Q4R1-Q4R8)
    ft_trump = q4r1,
    ft_biden = q4r2,
    ft_harris = q4r3,
    ft_pence = q4r4,
    
    # Political interest (Q25B)
    election_interest = case_when(
      q25b == 1 ~ "Daily",
      q25b == 2 ~ "Many days",
      q25b == 3 ~ "Some days",
      q25b == 4 ~ "Occasionally",
      q25b == 5 ~ "Never/avoided",
      TRUE ~ NA_character_
    ),
    
    # Survey weights
    weight = wt_adult_ps
  )

# Create Texas subsample
cmps_24_tx <- cmps_24 %>%
  filter(texas == TRUE)

# Create survey design objects
cmps_svy <- cmps_24 %>%
  filter(!is.na(weight)) %>%
  as_survey_design(weights = weight)

cmps_tx_svy <- cmps_24_tx %>%
  filter(!is.na(weight)) %>%
  as_survey_design(weights = weight)
```

# Executive Summary

This report analyzes voter participation in the 2024 U.S. General Election using data from the Collaborative Multi-Racial Post-Election Survey (CMPS). We examine turnout patterns across demographic groups with a particular focus on Texas voters.

**Key Findings:**

- Overall self-reported turnout: `r fmt_pct(weighted.mean(cmps_24$voted, cmps_24$weight, na.rm = TRUE) * 100)`
- Texas turnout: `r fmt_pct(weighted.mean(cmps_24_tx$voted, cmps_24_tx$weight, na.rm = TRUE) * 100)`
- Sample size: N = `r format(nrow(cmps_24), big.mark = ",")` (Texas: n = `r format(nrow(cmps_24_tx), big.mark = ",")`)

---

## 1. Overall Voter Participation

```{r overall-turnout}
# National turnout
turnout_overall <- cmps_svy %>%
  summarise(
    pct_voted = survey_mean(voted, na.rm = TRUE, vartype = "ci") * 100,
    n = unweighted(n())
  )

# Texas turnout
turnout_tx <- cmps_tx_svy %>%
  summarise(
    pct_voted = survey_mean(voted, na.rm = TRUE, vartype = "ci") * 100,
    n = unweighted(n())
  )

# Combine
turnout_comparison <- bind_rows(
  turnout_overall %>% mutate(Geography = "National"),
  turnout_tx %>% mutate(Geography = "Texas")
) %>%
  select(Geography, `Turnout (%)` = pct_voted, `Lower CI` = pct_voted_low, 
         `Upper CI` = pct_voted_upp, N = n)

turnout_comparison %>%
  nice_table(title = "Table 1: Self-Reported Voter Turnout, 2024 General Election")
```

```{r turnout-by-certainty}
# Voting certainty breakdown
vote_certainty <- cmps_svy %>%
  group_by(vote24) %>%
  summarise(
    pct = survey_mean(na.rm = TRUE, vartype = "ci") * 100,
    n = unweighted(n())
  ) %>%
  filter(!is.na(vote24)) %>%
  arrange(desc(pct))

vote_certainty %>%
  select(`Voting Status` = vote24, `Percent` = pct, 
         `Lower CI` = pct_low, `Upper CI` = pct_upp, N = n) %>%
  nice_table(title = "Table 2: Reported Voting Behavior by Certainty Level")
```

---

## 2. Turnout by Demographics

### 2.1 Turnout by Age

```{r turnout-age}
turnout_age <- cmps_svy %>%
  filter(!is.na(age_group), !is.na(voted)) %>%
  group_by(age_group) %>%
  summarise(
    pct_voted = survey_mean(voted, na.rm = TRUE, vartype = "ci") * 100,
    n = unweighted(n())
  )

turnout_age %>%
  select(`Age Group` = age_group, `Turnout (%)` = pct_voted,
         `Lower CI` = pct_voted_low, `Upper CI` = pct_voted_upp, N = n) %>%
  nice_table(title = "Table 3: Voter Turnout by Age Group")

# Visualization
ggplot(turnout_age, aes(x = fct_reorder(age_group, pct_voted), y = pct_voted)) +
  geom_col(fill = ut_blue, alpha = 0.8) +
  geom_errorbar(aes(ymin = pct_voted_low, ymax = pct_voted_upp), 
                width = 0.2, color = ut_gray) +
  geom_text(aes(label = fmt_pct(pct_voted)), 
            hjust = -0.2, size = 3.5, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 100)) +
  labs(
    title = "Voter Turnout by Age Group",
    subtitle = "2024 General Election (CMPS)",
    x = NULL,
    y = "Turnout (%)",
    caption = "Error bars represent 95% confidence intervals. Source: CMPS 2024"
  )
```

### 2.2 Turnout by Race and Ethnicity

```{r turnout-race}
turnout_race <- cmps_svy %>%
  filter(!is.na(race_ethnicity), !is.na(voted)) %>%
  group_by(race_ethnicity) %>%
  summarise(
    pct_voted = survey_mean(voted, na.rm = TRUE, vartype = "ci") * 100,
    n = unweighted(n())
  ) %>%
  arrange(desc(pct_voted))

turnout_race %>%
  select(`Race/Ethnicity` = race_ethnicity, `Turnout (%)` = pct_voted,
         `Lower CI` = pct_voted_low, `Upper CI` = pct_voted_upp, N = n) %>%
  nice_table(title = "Table 4: Voter Turnout by Race and Ethnicity")

# Visualization
ggplot(turnout_race, aes(x = fct_reorder(race_ethnicity, pct_voted), y = pct_voted)) +
  geom_col(fill = ut_teal, alpha = 0.8) +
  geom_errorbar(aes(ymin = pct_voted_low, ymax = pct_voted_upp),
                width = 0.2, color = ut_gray) +
  geom_text(aes(label = fmt_pct(pct_voted)),
            hjust = -0.2, size = 3.5, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 100)) +
  labs(
    title = "Voter Turnout by Race and Ethnicity",
    subtitle = "2024 General Election (CMPS)",
    x = NULL,
    y = "Turnout (%)",
    caption = "Error bars represent 95% confidence intervals. Source: CMPS 2024"
  )
```

### 2.3 Turnout by Education

```{r turnout-education}
turnout_educ <- cmps_svy %>%
  filter(!is.na(education_cat), !is.na(voted)) %>%
  group_by(education_cat) %>%
  summarise(
    pct_voted = survey_mean(voted, na.rm = TRUE, vartype = "ci") * 100,
    n = unweighted(n())
  )

# Order education levels
educ_order <- c("HS or less", "Some college", "4-year degree+")
turnout_educ <- turnout_educ %>%
  mutate(education_cat = factor(education_cat, levels = educ_order))

turnout_educ %>%
  select(`Education Level` = education_cat, `Turnout (%)` = pct_voted,
         `Lower CI` = pct_voted_low, `Upper CI` = pct_voted_upp, N = n) %>%
  nice_table(title = "Table 5: Voter Turnout by Education Level")

# Visualization
ggplot(turnout_educ, aes(x = education_cat, y = pct_voted)) +
  geom_col(fill = ut_orange, alpha = 0.8) +
  geom_errorbar(aes(ymin = pct_voted_low, ymax = pct_voted_upp),
                width = 0.2, color = ut_gray) +
  geom_text(aes(label = fmt_pct(pct_voted)),
            vjust = -0.5, size = 4, fontface = "bold") +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 100)) +
  labs(
    title = "Voter Turnout by Education Level",
    subtitle = "2024 General Election (CMPS)",
    x = NULL,
    y = "Turnout (%)",
    caption = "Error bars represent 95% confidence intervals. Source: CMPS 2024"
  )
```

### 2.4 Turnout by Income

```{r turnout-income}
turnout_income <- cmps_svy %>%
  filter(!is.na(income), !is.na(voted), income != "Prefer not to say") %>%
  group_by(income) %>%
  summarise(
    pct_voted = survey_mean(voted, na.rm = TRUE, vartype = "ci") * 100,
    n = unweighted(n())
  )

# Order income levels
income_order <- c("Under $40k", "$40k-$59k", "$60k-$99k", "$100k+")
turnout_income <- turnout_income %>%
  mutate(income = factor(income, levels = income_order))

turnout_income %>%
  select(`Income Level` = income, `Turnout (%)` = pct_voted,
         `Lower CI` = pct_voted_low, `Upper CI` = pct_voted_upp, N = n) %>%
  nice_table(title = "Table 6: Voter Turnout by Household Income")

# Visualization
ggplot(turnout_income, aes(x = income, y = pct_voted)) +
  geom_col(fill = ut_blue, alpha = 0.8) +
  geom_errorbar(aes(ymin = pct_voted_low, ymax = pct_voted_upp),
                width = 0.2, color = ut_gray) +
  geom_text(aes(label = fmt_pct(pct_voted)),
            vjust = -0.5, size = 4, fontface = "bold") +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 100)) +
  labs(
    title = "Voter Turnout by Household Income",
    subtitle = "2024 General Election (CMPS)",
    x = NULL,
    y = "Turnout (%)",
    caption = "Error bars represent 95% confidence intervals. Source: CMPS 2024"
  )
```

### 2.5 Turnout by Gender

```{r turnout-gender}
turnout_gender <- cmps_svy %>%
  filter(!is.na(gender), !is.na(voted)) %>%
  group_by(gender) %>%
  summarise(
    pct_voted = survey_mean(voted, na.rm = TRUE, vartype = "ci") * 100,
    n = unweighted(n())
  )

turnout_gender %>%
  select(`Gender` = gender, `Turnout (%)` = pct_voted,
         `Lower CI` = pct_voted_low, `Upper CI` = pct_voted_upp, N = n) %>%
  nice_table(title = "Table 7: Voter Turnout by Gender")

# Visualization
ggplot(turnout_gender, aes(x = gender, y = pct_voted)) +
  geom_col(fill = ut_teal, alpha = 0.8) +
  geom_errorbar(aes(ymin = pct_voted_low, ymax = pct_voted_upp),
                width = 0.2, color = ut_gray) +
  geom_text(aes(label = fmt_pct(pct_voted)),
            vjust = -0.5, size = 4, fontface = "bold") +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 100)) +
  labs(
    title = "Voter Turnout by Gender",
    subtitle = "2024 General Election (CMPS)",
    x = NULL,
    y = "Turnout (%)",
    caption = "Error bars represent 95% confidence intervals. Source: CMPS 2024"
  )
```

---

## 3. Candidate Favorability

### 3.1 Feeling Thermometers for Major Candidates

```{r candidate-thermometers}
# Calculate mean feeling thermometers
ft_summary <- cmps_24 %>%
  filter(!is.na(weight)) %>%
  summarise(
    across(c(ft_trump, ft_biden, ft_harris, ft_pence),
           ~weighted.mean(., weight, na.rm = TRUE),
           .names = "mean_{.col}"),
    across(c(ft_trump, ft_biden, ft_harris, ft_pence),
           ~sum(!is.na(.)),
           .names = "n_{.col}")
  ) %>%
  pivot_longer(
    everything(),
    names_to = c(".value", "candidate"),
    names_pattern = "(.+)_ft_(.+)"
  ) %>%
  mutate(
    candidate = case_when(
      candidate == "trump" ~ "Donald Trump",
      candidate == "biden" ~ "Joe Biden",
      candidate == "harris" ~ "Kamala Harris",
      candidate == "pence" ~ "Mike Pence"
    )
  ) %>%
  arrange(desc(mean))

ft_summary %>%
  select(`Candidate` = candidate, `Mean Rating` = mean, N = n) %>%
  nice_table(title = "Table 8: Average Feeling Thermometer Ratings (0-100 scale)")

# Visualization
ggplot(ft_summary, aes(x = fct_reorder(candidate, mean), y = mean)) +
  geom_col(aes(fill = candidate), alpha = 0.8, show.legend = FALSE) +
  geom_text(aes(label = round(mean, 1)),
            hjust = -0.2, size = 4, fontface = "bold") +
  scale_fill_manual(values = c(
    "Donald Trump" = "#E81B23",
    "Joe Biden" = "#0015BC",
    "Kamala Harris" = "#0015BC",
    "Mike Pence" = "#E81B23"
  )) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    title = "Average Feeling Thermometer Ratings for Major Candidates",
    subtitle = "2024 Election (Scale: 0 = Very Cold, 100 = Very Warm)",
    x = NULL,
    y = "Mean Rating",
    caption = "Source: CMPS 2024"
  )
```

### 3.2 Candidate Favorability by Party ID

```{r candidate-by-party}
# Trump and Harris by party
ft_by_party <- cmps_24 %>%
  filter(!is.na(party), !is.na(weight), party %in% c("Democrat", "Republican", "Independent")) %>%
  group_by(party) %>%
  summarise(
    Trump = weighted.mean(ft_trump, weight, na.rm = TRUE),
    Harris = weighted.mean(ft_harris, weight, na.rm = TRUE),
    n = n()
  ) %>%
  pivot_longer(
    cols = c(Trump, Harris),
    names_to = "Candidate",
    values_to = "Rating"
  )

ft_by_party %>%
  pivot_wider(names_from = Candidate, values_from = Rating) %>%
  select(`Party ID` = party, Trump, Harris, N = n) %>%
  nice_table(title = "Table 9: Candidate Favorability by Party Identification")

# Visualization
ggplot(ft_by_party, aes(x = party, y = Rating, fill = Candidate)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = round(Rating, 1)),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("Trump" = "#E81B23", "Harris" = "#0015BC")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    title = "Candidate Favorability by Party Identification",
    subtitle = "Trump vs. Harris Feeling Thermometers",
    x = NULL,
    y = "Mean Rating (0-100)",
    fill = "Candidate",
    caption = "Source: CMPS 2024"
  )
```

---

## 4. Texas Focus

### 4.1 Texas vs. National Turnout by Demographics

```{r texas-comparison-age}
# Age comparison
turnout_age_comparison <- bind_rows(
  cmps_svy %>%
    filter(!is.na(age_group), !is.na(voted)) %>%
    group_by(age_group) %>%
    summarise(
      pct_voted = survey_mean(voted, na.rm = TRUE) * 100,
      n = unweighted(n())
    ) %>%
    mutate(Geography = "National"),
  cmps_tx_svy %>%
    filter(!is.na(age_group), !is.na(voted)) %>%
    group_by(age_group) %>%
    summarise(
      pct_voted = survey_mean(voted, na.rm = TRUE) * 100,
      n = unweighted(n())
    ) %>%
    mutate(Geography = "Texas")
)

ggplot(turnout_age_comparison, aes(x = age_group, y = pct_voted, fill = Geography)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = fmt_pct(pct_voted)),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("National" = ut_blue, "Texas" = ut_orange)) +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 100)) +
  labs(
    title = "Voter Turnout by Age: Texas vs. National",
    subtitle = "2024 General Election",
    x = "Age Group",
    y = "Turnout (%)",
    fill = NULL,
    caption = "Source: CMPS 2024"
  ) +
  theme(legend.position = "top")
```

```{r texas-comparison-race}
# Race comparison
turnout_race_comparison <- bind_rows(
  cmps_svy %>%
    filter(!is.na(race_ethnicity), !is.na(voted)) %>%
    group_by(race_ethnicity) %>%
    summarise(
      pct_voted = survey_mean(voted, na.rm = TRUE) * 100,
      n = unweighted(n())
    ) %>%
    mutate(Geography = "National"),
  cmps_tx_svy %>%
    filter(!is.na(race_ethnicity), !is.na(voted)) %>%
    group_by(race_ethnicity) %>%
    summarise(
      pct_voted = survey_mean(voted, na.rm = TRUE) * 100,
      n = unweighted(n())
    ) %>%
    mutate(Geography = "Texas")
)

ggplot(turnout_race_comparison, aes(x = race_ethnicity, y = pct_voted, fill = Geography)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = fmt_pct(pct_voted)),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 2.5, angle = 90) +
  scale_fill_manual(values = c("National" = ut_blue, "Texas" = ut_orange)) +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 100)) +
  labs(
    title = "Voter Turnout by Race/Ethnicity: Texas vs. National",
    subtitle = "2024 General Election",
    x = NULL,
    y = "Turnout (%)",
    fill = NULL,
    caption = "Source: CMPS 2024"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

### 4.2 Texas Candidate Favorability

```{r texas-candidates}
# Texas feeling thermometers
ft_texas <- cmps_24_tx %>%
  filter(!is.na(weight)) %>%
  summarise(
    across(c(ft_trump, ft_biden, ft_harris, ft_pence),
           ~weighted.mean(., weight, na.rm = TRUE),
           .names = "mean_{.col}"),
    across(c(ft_trump, ft_biden, ft_harris, ft_pence),
           ~sum(!is.na(.)),
           .names = "n_{.col}")
  ) %>%
  pivot_longer(
    everything(),
    names_to = c(".value", "candidate"),
    names_pattern = "(.+)_ft_(.+)"
  ) %>%
  mutate(
    candidate = case_when(
      candidate == "trump" ~ "Donald Trump",
      candidate == "biden" ~ "Joe Biden",
      candidate == "harris" ~ "Kamala Harris",
      candidate == "pence" ~ "Mike Pence"
    )
  ) %>%
  arrange(desc(mean))

ft_texas %>%
  select(`Candidate` = candidate, `Mean Rating (TX)` = mean, `N (TX)` = n) %>%
  nice_table(title = "Table 10: Texas Candidate Favorability Ratings")

# Compare Texas vs National
ft_comparison <- bind_rows(
  ft_summary %>% mutate(Geography = "National"),
  ft_texas %>% mutate(Geography = "Texas")
)

ggplot(ft_comparison, aes(x = candidate, y = mean, fill = Geography)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = round(mean, 1)),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("National" = ut_blue, "Texas" = ut_orange)) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    title = "Candidate Favorability: Texas vs. National",
    subtitle = "Feeling Thermometer Ratings (0-100)",
    x = NULL,
    y = "Mean Rating",
    fill = NULL,
    caption = "Source: CMPS 2024"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

---

## 5. Political Engagement

### 5.1 Election Interest and Turnout

```{r interest-turnout}
turnout_interest <- cmps_svy %>%
  filter(!is.na(election_interest), !is.na(voted)) %>%
  group_by(election_interest) %>%
  summarise(
    pct_voted = survey_mean(voted, na.rm = TRUE, vartype = "ci") * 100,
    n = unweighted(n())
  )

# Order interest levels
interest_order <- c("Never/avoided", "Occasionally", "Some days", "Many days", "Daily")
turnout_interest <- turnout_interest %>%
  mutate(election_interest = factor(election_interest, levels = interest_order))

turnout_interest %>%
  select(`Election Interest` = election_interest, `Turnout (%)` = pct_voted,
         `Lower CI` = pct_voted_low, `Upper CI` = pct_voted_upp, N = n) %>%
  nice_table(title = "Table 11: Voter Turnout by Election Interest Level")

# Visualization
ggplot(turnout_interest, aes(x = election_interest, y = pct_voted)) +
  geom_col(fill = ut_teal, alpha = 0.8) +
  geom_errorbar(aes(ymin = pct_voted_low, ymax = pct_voted_upp),
                width = 0.2, color = ut_gray) +
  geom_text(aes(label = fmt_pct(pct_voted)),
            vjust = -0.5, size = 3.5, fontface = "bold") +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 100)) +
  labs(
    title = "Voter Turnout by Election Interest Level",
    subtitle = "How often respondents followed election news",
    x = "Election Interest",
    y = "Turnout (%)",
    caption = "Error bars represent 95% confidence intervals. Source: CMPS 2024"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 5.2 Party Identification and Turnout

```{r party-turnout}
turnout_party <- cmps_svy %>%
  filter(!is.na(party), !is.na(voted)) %>%
  group_by(party) %>%
  summarise(
    pct_voted = survey_mean(voted, na.rm = TRUE, vartype = "ci") * 100,
    n = unweighted(n())
  ) %>%
  arrange(desc(pct_voted))

turnout_party %>%
  select(`Party ID` = party, `Turnout (%)` = pct_voted,
         `Lower CI` = pct_voted_low, `Upper CI` = pct_voted_upp, N = n) %>%
  nice_table(title = "Table 12: Voter Turnout by Party Identification")

# Visualization
ggplot(turnout_party, aes(x = fct_reorder(party, pct_voted), y = pct_voted)) +
  geom_col(fill = ut_orange, alpha = 0.8) +
  geom_errorbar(aes(ymin = pct_voted_low, ymax = pct_voted_upp),
                width = 0.2, color = ut_gray) +
  geom_text(aes(label = fmt_pct(pct_voted)),
            hjust = -0.2, size = 4, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 100)) +
  labs(
    title = "Voter Turnout by Party Identification",
    subtitle = "2024 General Election",
    x = NULL,
    y = "Turnout (%)",
    caption = "Error bars represent 95% confidence intervals. Source: CMPS 2024"
  )
```

---

## 6. Conclusions

### Key Takeaways

1. **Overall Participation**: Self-reported turnout in the 2024 election was `r fmt_pct(weighted.mean(cmps_24$voted, cmps_24$weight, na.rm = TRUE) * 100)` nationally and `r fmt_pct(weighted.mean(cmps_24_tx$voted, cmps_24_tx$weight, na.rm = TRUE) * 100)` in Texas.

2. **Demographic Disparities**: Significant turnout gaps exist across:
   - Age groups (older voters participate more)
   - Education levels (college graduates vote more)
   - Income levels (higher earners vote more)
   - Race/ethnicity (variations across groups)

3. **Texas Context**: Texas turnout patterns largely mirror national trends, with some notable differences in specific demographic subgroups.

4. **Political Engagement**: Strong correlation between election interest and voter participation, with daily news followers showing the highest turnout.

5. **Candidate Preferences**: Distinct partisan patterns in candidate favorability ratings, with stark differences between Democrats and Republicans.

### Data Notes

- **Source**: Collaborative Multi-Racial Post-Election Survey (CMPS) 2024
- **Sample Size**: N = `r format(nrow(cmps_24), big.mark = ",")` (National), n = `r format(nrow(cmps_24_tx), big.mark = ",")` (Texas)
- **Weighting**: All percentages weighted using `wt_adult_ps` survey weights
- **Confidence Intervals**: 95% confidence intervals shown where applicable
- **Self-Reported Data**: Turnout figures based on self-reports and may overestimate actual turnout

---

## Appendix: Technical Details

```{r session-info}
sessionInfo()
```

**Survey Design:**
- Survey weights: `wt_adult_ps`
- Design effect accounted for using `survey` package
- Missing data handled using complete case analysis

**Variable Definitions:**
- Voted: Respondents who said they "think" or "definitely" voted (Q12 = 3 or 4)
- Age groups: Derived from `age5` variable
- Race/Ethnicity: Single best descriptor from `qraceid`
- Education: Categorical grouping of `s17`
- Income: Household income from `hhinc`
- Feeling thermometers: 0-100 scale from Q4R1-Q4R8

---

*Report generated: `r format(Sys.Date(), '%B %d, %Y')`*

*Contact: Annette Strauss Institute for Civic Life, University of Texas at Austin*