---
title: "Voter Apathy and Non-Participation in the 2024 U.S. General Election"
subtitle: "An Exploratory Analysis with Focus on Texas"
author: "Annette Strauss Institute for Civic Life, University of Texas at Austin"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: hide
    fig_width: 11
    fig_height: 7
    df_print: paged
---

<style>
.main-container {
  max-width: 1400px;
}
body {
  font-size: 11pt;
}
h1 {
  color: #BF5700;
  border-bottom: 2px solid #BF5700;
  padding-bottom: 10px;
}
h2 {
  color: #005F86;
}
.caption {
  font-style: italic;
  color: #666;
  font-size: 0.9em;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.align = "center",
  cache = FALSE
)

# Load packages
library(tidyverse)
library(haven)
library(labelled)
library(survey)
library(scales)
library(knitr)
library(kableExtra)
library(patchwork)
library(forcats)

# Set UT Austin branding colors
ut_orange <- "#BF5700"
ut_blue <- "#005F86"
ut_teal <- "#00A9B7"
ut_gray <- "#666666"

# Custom ggplot theme
theme_asi <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 13, color = ut_blue),
      plot.subtitle = element_text(size = 10, color = ut_gray, margin = margin(b = 10)),
      plot.caption = element_text(size = 8, color = ut_gray, hjust = 0),
      axis.title = element_text(size = 10, face = "bold"),
      legend.position = "bottom",
      legend.title = element_text(face = "bold", size = 9),
      strip.text = element_text(face = "bold", size = 10, color = ut_blue),
      strip.background = element_rect(fill = "gray95", color = NA),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "gray80", fill = NA)
    )
}

theme_set(theme_asi())

# Color palette for participation
part_colors <- c(
  "Voted" = "#2ecc71",
  "Registered, no vote" = "#f39c12", 
  "Not registered" = "#e74c3c"
)

# Helper: Create publication-ready tables with sample sizes
nice_table <- function(df, title = NULL, digits = 2, include_n = TRUE) {
  raw_names <- names(df)
  pretty_names <- raw_names %>%
    str_replace_all("_", " ") %>%
    str_replace("^pct ", "") %>%
    str_replace("^pct", "") %>%
    str_replace("^n ", "N ")
  pct_cols <- str_detect(raw_names, "^pct[_ ]")
  pretty_names[pct_cols] <- paste0(pretty_names[pct_cols], " (%)")
  pretty_names[str_detect(raw_names, "^n[_ ]")] <- str_replace(pretty_names[str_detect(raw_names, "^n[_ ]")], "^n ", "N ")

  df %>%
    mutate(across(where(is.numeric), ~ round(., digits = digits))) %>%
    kable(
      caption = title, 
      col.names = pretty_names,
      format = "html",
      booktabs = TRUE,
      align = "c"
    ) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE,
      position = "left",
      font_size = 11
    ) %>%
    row_spec(0, bold = TRUE, color = "white", background = ut_blue)
}

# Helper: Format percentages
fmt_pct <- function(x, digits = 1) {
  paste0(round(x * 100, digits), "%")
}
```

```{r data-precompute, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
# Early data load so inline values in the Executive Summary are data-driven
anes_path <- "raw_data/anes.csv"
ces_path <- "raw_data/ces.dta"

anes_raw <- read_csv(anes_path, show_col_types = FALSE)
ces_raw <- read_dta(ces_path)

# ANES prep
anes_24 <- anes_raw %>%
  filter(VCF0004 == 2024) %>%
  mutate(
    state = as_factor(VCF0901b),
    state_chr = tolower(as.character(state)),
    texas = if_else(is.na(state_chr), FALSE, str_detect(state_chr, "texas|tx|tx -"))
  ) %>%
  mutate(
    participation = case_when(
      VCF0703 == 1 ~ "Not registered",
      VCF0703 == 2 ~ "Registered, no vote",
      VCF0703 == 3 ~ "Voted",
      TRUE ~ NA_character_
    ),
    participation = factor(participation, levels = c("Voted", "Registered, no vote", "Not registered")),
    voted = VCF0703 == 3,
    registered = VCF0703 %in% c(2, 3),
    age = VCF0101,
    age_group = case_when(
      age < 30 ~ "18-29",
      age < 45 ~ "30-44",
      age < 65 ~ "45-64",
      age >= 65 ~ "65+",
      TRUE ~ NA_character_
    ),
    gender = case_when(
      VCF0104 == 1 ~ "Male",
      VCF0104 == 2 ~ "Female",
      VCF0104 == 3 ~ "Other",
      TRUE ~ NA_character_
    ),
    race = case_when(
      VCF0105a == 1 ~ "White",
      VCF0105a == 2 ~ "Black",
      VCF0105a == 3 ~ "Hispanic",
      VCF0105a == 4 ~ "Asian/PI",
      VCF0105a %in% c(5, 6, 7) ~ "Other/Mixed",
      TRUE ~ NA_character_
    ),
    educ = case_when(
      VCF0110 == 1 ~ "< High School",
      VCF0110 == 2 ~ "High School",
      VCF0110 == 3 ~ "Some College",
      VCF0110 == 4 ~ "College+",
      TRUE ~ NA_character_
    ),
    educ = factor(educ, levels = c("< High School", "High School", "Some College", "College+")),
    efficacy = case_when(
      VCF0609 == 1 ~ "Agree strongly",
      VCF0609 == 2 ~ "Agree somewhat",
      VCF0609 == 3 ~ "Neither",
      VCF0609 == 4 ~ "Disagree somewhat",
      VCF0609 == 5 ~ "Disagree strongly",
      TRUE ~ NA_character_
    ),
    efficacy = factor(efficacy, levels = c("Agree strongly", "Agree somewhat", "Neither", "Disagree somewhat", "Disagree strongly")),
    see_party_diff = case_when(
      VCF0501 == 1 ~ "Major differences",
      VCF0501 == 2 ~ "Not much difference",
      TRUE ~ NA_character_
    ),
    interest = case_when(
      VCF0310 == 1 ~ "Very interested",
      VCF0310 == 2 ~ "Somewhat interested",
      VCF0310 == 3 ~ "Not much interested",
      TRUE ~ NA_character_
    ),
    interest = factor(interest, levels = c("Very interested", "Somewhat interested", "Not much interested")),
    see_party_diff = factor(see_party_diff, levels = c("Major differences", "Not much difference")),
    weight = VCF0009z
  ) %>%
  filter(!is.na(participation))

# CES prep
ces_24 <- ces_raw %>%
  filter(year == 2024) %>%
  mutate(
    state_label = as_factor(st),
    state = as.character(state_label),
    texas = state %in% c("TX", "Texas"),
    voted_validated = case_when(
      vv_turnout_gvm == 1 ~ TRUE,
      vv_turnout_gvm == 2 ~ FALSE,
      TRUE ~ NA
    ),
    voted_self = case_when(
      voted_turnout_self %in% c(1) ~ TRUE,
      voted_turnout_self %in% c(0, 2) ~ FALSE,
      TRUE ~ NA
    ),
    reg_status = case_when(
      vv_regstatus == 1 ~ "Active",
      vv_regstatus == 2 ~ "Inactive",
      vv_regstatus == 3 ~ "Unregistered",
      vv_regstatus == 4 ~ "Matched, dropped/purged",
      vv_regstatus == 5 ~ "Unmatched",
      TRUE ~ NA_character_
    ),
    registered_validated = case_when(
      vv_regstatus %in% c(1, 2) ~ TRUE,
      vv_regstatus == 3 ~ FALSE,
      TRUE ~ NA
    ),
    weight_full = weight,
    weight_validated = if_else(!is.na(vvweight), vvweight, weight),
    participation = case_when(
      !is.na(voted_validated) & voted_validated ~ "Voted",
      !is.na(voted_validated) & !voted_validated & registered_validated ~ "Registered, no vote",
      !is.na(voted_validated) & registered_validated == FALSE ~ "Not registered",
      is.na(voted_validated) & !is.na(voted_self) & voted_self ~ "Voted",
      is.na(voted_validated) & !is.na(voted_self) & !voted_self ~ "Registered, no vote",
      TRUE ~ NA_character_
    ),
    participation = factor(participation, levels = c("Voted", "Registered, no vote", "Not registered")),
    has_validated = !is.na(voted_validated) & !is.na(voted_self),
    over_reports = voted_self == TRUE & voted_validated == FALSE,
    under_reports = voted_self == FALSE & voted_validated == TRUE,
    accurate_report = voted_self == voted_validated,
    age_group = case_when(
      age < 30 ~ "18-29",
      age < 45 ~ "30-44",
      age < 65 ~ "45-64",
      age >= 65 ~ "65+",
      TRUE ~ NA_character_
    ),
    age_group = factor(age_group, levels = c("18-29", "30-44", "45-64", "65+")),
    gender_cat = as_factor(gender),
    race_cat = as_factor(race),
    educ_cat = as_factor(educ),
    ideology = case_when(
      ideo5 %in% 1:2 ~ "Liberal",
      ideo5 == 3 ~ "Moderate",
      ideo5 %in% 4:5 ~ "Conservative",
      TRUE ~ NA_character_
    ),
    pid = as_factor(pid3),
    news_interest = case_when(
      newsint == 1 ~ "Most of the time",
      newsint == 2 ~ "Some of the time",
      newsint == 3 ~ "Only now and then",
      newsint == 4 ~ "Hardly at all",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(participation))

# Summary metrics for inline text
anes_turnout_by_educ <- anes_24 %>%
  filter(!is.na(educ)) %>%
  group_by(educ) %>%
  summarise(pct = weighted.mean(voted, weight, na.rm = TRUE) * 100, .groups = "drop")

educ_low <- anes_turnout_by_educ$pct[anes_turnout_by_educ$educ == "< High School"]
educ_high <- anes_turnout_by_educ$pct[anes_turnout_by_educ$educ == "College+"]
educ_gap <- educ_high - educ_low
educ_ratio <- educ_high / educ_low

ces_turnout_gap <- ces_24 %>%
  filter(has_validated) %>%
  group_by(texas) %>%
  summarise(pct = weighted.mean(voted_validated, weight_validated, na.rm = TRUE) * 100, .groups = "drop") %>%
  summarise(gap = pct[texas == FALSE] - pct[texas == TRUE]) %>% pull(gap)

anes_turnout_gap <- anes_24 %>%
  group_by(texas) %>%
  summarise(pct = weighted.mean(voted, weight, na.rm = TRUE) * 100, .groups = "drop") %>%
  summarise(gap = pct[texas == FALSE] - pct[texas == TRUE]) %>% pull(gap)

over_report_rate <- ces_24 %>%
  filter(has_validated) %>%
  summarise(rate = weighted.mean(over_reports, weight_validated, na.rm = TRUE) * 100) %>% pull(rate)

eff_turnout <- anes_24 %>%
  filter(!is.na(efficacy)) %>%
  group_by(efficacy) %>%
  summarise(pct = weighted.mean(voted, weight, na.rm = TRUE) * 100, .groups = "drop")
eff_low <- eff_turnout$pct[eff_turnout$efficacy == "Agree strongly"]
eff_high <- eff_turnout$pct[eff_turnout$efficacy == "Disagree strongly"]
eff_gap <- eff_high - eff_low

tx_young_low <- ces_24 %>%
  filter(texas, has_validated, age_group == "18-29", educ_cat %in% c("No HS", "High School Graduate")) %>%
  summarise(pct_not_reg = weighted.mean(!registered_validated, weight_validated, na.rm = TRUE) * 100, n = n())
tx_young_low_rate <- tx_young_low$pct_not_reg
tx_young_low_n <- tx_young_low$n

hispanic_tx_nonreg <- ces_24 %>%
  filter(texas, has_validated, race_cat == "Hispanic", educ_cat == "No HS") %>%
  summarise(rate = weighted.mean(!registered_validated, weight_validated, na.rm = TRUE) * 100, n = n())
hispanic_tx_nonreg_rate <- hispanic_tx_nonreg$rate
hispanic_tx_nonreg_n <- hispanic_tx_nonreg$n

tx_nonreg_young <- ces_24 %>%
  filter(texas, has_validated, age_group == "18-29") %>%
  summarise(rate = weighted.mean(!registered_validated, weight_validated, na.rm = TRUE) * 100) %>%
  pull(rate)
tx_nonreg_older <- ces_24 %>%
  filter(texas, has_validated, age_group != "18-29") %>%
  summarise(rate = weighted.mean(!registered_validated, weight_validated, na.rm = TRUE) * 100) %>%
  pull(rate)
tx_young_gap <- tx_nonreg_young - tx_nonreg_older

interest_turnout <- anes_24 %>%
  filter(!is.na(interest)) %>%
  group_by(interest) %>%
  summarise(pct = weighted.mean(voted, weight, na.rm = TRUE) * 100, .groups = "drop")
interest_gap <- interest_turnout$pct[interest_turnout$interest == "Very interested"] -
  interest_turnout$pct[interest_turnout$interest == "Not much interested"]

party_turnout <- anes_24 %>%
  filter(!is.na(see_party_diff)) %>%
  group_by(see_party_diff) %>%
  summarise(pct = weighted.mean(voted, weight, na.rm = TRUE) * 100, .groups = "drop")
party_gap <- party_turnout$pct[party_turnout$see_party_diff == "Major differences"] -
  party_turnout$pct[party_turnout$see_party_diff == "Not much difference"]

apathy_low <- anes_24 %>%
  filter(efficacy == "Agree strongly", interest == "Not much interested") %>%
  summarise(pct = weighted.mean(voted, weight, na.rm = TRUE) * 100, n = n())
apathy_high <- anes_24 %>%
  filter(efficacy == "Disagree strongly", interest == "Very interested") %>%
  summarise(pct = weighted.mean(voted, weight, na.rm = TRUE) * 100, n = n())
apathy_gap <- apathy_high$pct - apathy_low$pct

black_tx_registered_no_vote <- ces_24 %>%
  filter(texas, has_validated, race_cat == "Black") %>%
  summarise(rate = weighted.mean(participation == "Registered, no vote", weight_validated, na.rm = TRUE) * 100, n = n())
black_tx_no_vote_rate <- black_tx_registered_no_vote$rate
black_tx_no_vote_n <- black_tx_registered_no_vote$n

tx_effect_model <- ces_24 %>%
  filter(has_validated) %>%
  mutate(voted_num = as.numeric(voted_validated)) %>%
  lm(voted_num ~ texas + age_group + race_cat + educ_cat + gender_cat, data = ., weights = weight_validated)
texas_effect_demog <- coef(tx_effect_model)["texasTRUE"] * 100
```

# Executive Summary {-}

This exploratory analysis examines **voter apathy and non-participation** in the 2024 U.S. General Election using two nationally representative surveys: the American National Election Studies (ANES) and the Cooperative Election Study (CES). As researchers at the **Annette Strauss Institute for Civic Life** based in Texas, we give special attention to Texas-national comparisons and the unique challenges Texas voters face.

**Key Findings:**

- **The education gap persists:** College+ turnout `r sprintf('%.1f%%', educ_high)` vs. `<HS` `r sprintf('%.1f%%', educ_low)` (gap `r sprintf('%.1f pts', educ_gap)`)
- **Texas lags behind:** Turnout gap vs. nation: CES (validated) `r sprintf('%.1f pts', ces_turnout_gap)`, ANES `r sprintf('%.1f pts', anes_turnout_gap)`
- **Over-reporting is common:** `r sprintf('%.1f%%', over_report_rate)` self-report voting while validated as non-voters
- **Political efficacy matters most:** High- vs. low-efficacy turnout gap `r sprintf('%.1f pts', eff_gap)`
- **Young + low-SES Texans face compounded barriers:** Non-registration among TX age 18-29, ≤HS `r sprintf('%.1f%% (n=%s)', tx_young_low_rate, format(tx_young_low_n, big.mark=","))`

---

# Introduction

## Research Questions

This study examines voter apathy and barriers to political participation in the 2024 U.S. General Election, with focused attention on Texas. We investigate:

1. **Who doesn't register or vote?** Demographic and attitudinal patterns of non-participation
2. **Why do people not register?** Structural and psychological barriers to registration
3. **Why do registered voters abstain?** Predictors of "registered, no vote" status
4. **How does Texas compare to the nation?** State-specific patterns and disparities
5. **Who over-reports voting?** Analyzing self-reported vs. validated turnout (social desirability bias)
6. **Intersectionality:** How do demographic factors compound to create participation gaps?

## Data Sources

**ANES Time Series Cumulative Data File (1948-2024)**

- **2024 Sample:** N = 5,521 respondents
- **Texas subsample:** N = `r format(sum(anes_24$texas, na.rm=TRUE), big.mark=",")` (unweighted)
- **Mode:** Combined sample (web, face-to-face, video, phone)
- **Key feature:** Registration and turnout summary variable (VCF0703)
- **Weight:** VCF0009z (full combined sample, Type 0 variables for 1970)

**CES Cumulative Common Content (2006-2024)**

- **2024 Sample:** N = 60,000 respondents
- **Texas subsample:** N = `r format(sum(ces_24$texas, na.rm=TRUE), big.mark=",")` (unweighted)
- **Mode:** Online survey (YouGov panel)
- **Key feature:** Validated voter file matching (vv_turnout_gvm, vv_regstatus)
- **Weight:** `weight` (year-specific) or `vvweight` (validated registered voters only)

## Survey Design Considerations

**Weights:**
- All descriptive statistics use appropriate survey weights
- ANES: VCF0009z accounts for multi-mode design and post-stratification
- CES: `weight` for full sample; `vvweight` when restricting to validated registered voters

**Validated Turnout:**
- CES matches respondents to commercial voter files (L2, Catalist, etc.)
- Allows detection of over-reporting (claiming to vote when records show otherwise)
- Not all respondents successfully matched; missing values possible

**Missing Data:**
- Listwise deletion for complete-case analysis
- Missing values explicitly coded; no imputation
- Sample size (N) reported in all tables and figures

**Timing of Measures:**
- ANES: Pre-election attitudes measured before November 2024; post-election turnout after
- CES: Pre-wave (October) and post-wave (November-December); validated turnout updated months later

---

# Data Preparation

## Load Raw Data

```{r load-data}
# Define paths
anes_path <- "raw_data/anes.csv"
ces_path <- "raw_data/ces.dta"

# Load ANES
cat("Loading ANES Time Series Cumulative Data File...\n")
anes_raw <- read_csv(anes_path, show_col_types = FALSE)
cat(sprintf("  Total records: %s rows × %s variables\n", 
            format(nrow(anes_raw), big.mark = ","), 
            format(ncol(anes_raw), big.mark = ",")))

# Load CES
cat("\nLoading CES Cumulative Common Content...\n")
ces_raw <- read_dta(ces_path)
cat(sprintf("  Total records: %s rows × %s variables\n", 
            format(nrow(ces_raw), big.mark = ","), 
            format(ncol(ces_raw), big.mark = ",")))

# Filter to 2024
anes_2024_n <- nrow(filter(anes_raw, VCF0004 == 2024))
ces_2024_n <- nrow(filter(ces_raw, year == 2024))

cat(sprintf("\n2024 Samples:\n"))
cat(sprintf("  ANES 2024: %s respondents\n", format(anes_2024_n, big.mark = ",")))
cat(sprintf("  CES 2024: %s respondents\n", format(ces_2024_n, big.mark = ",")))
```

## ANES 2024: Variable Selection and Recoding

```{r prepare-anes, cache=TRUE}
# Key variables from ANES codebook:
# VCF0703: Registration & Turnout Summary (1=Not reg, 2=Reg no vote, 3=Voted)
# VCF0901a: State FIPS code
# VCF0901b: State postal abbreviation
# VCF0009z: Full sample weight
# Demographics: VCF0101 (age), VCF0104 (gender), VCF0105a (race), VCF0110 (educ)
# Attitudes: VCF0609, VCF0605, VCF0310, VCF0501, VCF0803

anes_24 <- anes_raw %>%
  filter(VCF0004 == 2024) %>%
  select(
    # ID and geography
    VCF0006, VCF0901a, VCF0901b,
    # Weights
    VCF0009z,
    # Turnout/Registration
    VCF0703, VCF0701, VCF0702,
    # Demographics
    VCF0101, VCF0104, VCF0105a, VCF0105b, VCF0110, VCF0114, VCF0140a, VCF0147,
    # Attitudes
    VCF0301, VCF0609, VCF0605, VCF0310, VCF0803, VCF0450, VCF0501, VCF0311, VCF0613
  ) %>%
  mutate(
    # Geography
    state_fips = VCF0901a,
    state = as_factor(VCF0901b),
    state_chr = tolower(as.character(state)),
    texas = if_else(is.na(state_chr), FALSE, str_detect(state_chr, "texas|tx|tx -")),
    
    # Outcome: Participation
    participation = case_when(
      VCF0703 == 1 ~ "Not registered",
      VCF0703 == 2 ~ "Registered, no vote",
      VCF0703 == 3 ~ "Voted",
      TRUE ~ NA_character_
    ),
    participation = factor(
      participation, 
      levels = c("Voted", "Registered, no vote", "Not registered")
    ),
    
    # Binary: Voted vs. Not
    voted = (VCF0703 == 3),
    registered = (VCF0703 %in% c(2, 3)),
    
    # Demographics
    age = VCF0101,
    age_group = case_when(
      age < 30 ~ "18-29",
      age < 45 ~ "30-44",
      age < 65 ~ "45-64",
      age >= 65 ~ "65+",
      TRUE ~ NA_character_
    ),
    age_group = factor(age_group, levels = c("18-29", "30-44", "45-64", "65+")),
    
    gender = case_when(
      VCF0104 == 1 ~ "Male",
      VCF0104 == 2 ~ "Female",
      VCF0104 == 3 ~ "Other",
      TRUE ~ NA_character_
    ),
    
    race = case_when(
      VCF0105a == 1 ~ "White",
      VCF0105a == 2 ~ "Black",
      VCF0105a == 3 ~ "Hispanic",
      VCF0105a == 4 ~ "Asian/PI",
      VCF0105a %in% c(5, 6, 7) ~ "Other/Mixed",
      TRUE ~ NA_character_
    ),
    
    race_simple = case_when(
      VCF0105b == 1 ~ "White",
      VCF0105b == 2 ~ "Black",
      VCF0105b == 3 ~ "Hispanic",
      VCF0105b == 4 ~ "Other",
      TRUE ~ NA_character_
    ),
    
    educ = case_when(
      VCF0110 == 1 ~ "< High School",
      VCF0110 == 2 ~ "High School",
      VCF0110 == 3 ~ "Some College",
      VCF0110 == 4 ~ "College+",
      TRUE ~ NA_character_
    ),
    educ = factor(educ, levels = c("< High School", "High School", "Some College", "College+")),
    
    educ_detailed = case_when(
      VCF0140a == 1 ~ "< 9th grade",
      VCF0140a == 2 ~ "9-12th, no diploma",
      VCF0140a == 3 ~ "High school graduate",
      VCF0140a == 4 ~ "Some college, no degree",
      VCF0140a == 5 ~ "Associate degree",
      VCF0140a == 6 ~ "Bachelor's degree",
      VCF0140a == 7 ~ "Graduate degree",
      TRUE ~ NA_character_
    ),
    
    income = VCF0114,
    
    # Attitudes
    # VCF0609: "Government officials don't care what people think"
    # 1=Agree strongly (LOW efficacy), 5=Disagree strongly (HIGH efficacy)
    efficacy_raw = VCF0609,
    efficacy = case_when(
      VCF0609 == 1 ~ "Agree strongly",
      VCF0609 == 2 ~ "Agree somewhat",
      VCF0609 == 3 ~ "Neither",
      VCF0609 == 4 ~ "Disagree somewhat",
      VCF0609 == 5 ~ "Disagree strongly",
      TRUE ~ NA_character_
    ),
    efficacy = factor(efficacy, levels = c("Agree strongly", "Agree somewhat", "Neither", "Disagree somewhat", "Disagree strongly")),
    
    # VCF0605: "Federal govt run by few interests (1) or for benefit of all (5)"
    trust_govt_raw = VCF0605,
    trust_govt = case_when(
      VCF0605 %in% 1:2 ~ "Low",
      VCF0605 == 3 ~ "Medium",
      VCF0605 %in% 4:5 ~ "High",
      TRUE ~ NA_character_
    ),
    trust_govt = factor(trust_govt, levels = c("Low", "Medium", "High")),
    
    # VCF0310: Interest in elections (1=Very, 2=Somewhat, 3=Not much)
    interest_raw = VCF0310,
    interest = case_when(
      VCF0310 == 1 ~ "Very interested",
      VCF0310 == 2 ~ "Somewhat interested",
      VCF0310 == 3 ~ "Not much interested",
      TRUE ~ NA_character_
    ),
    interest = factor(interest, levels = c("Very interested", "Somewhat interested", "Not much interested")),
    
    # VCF0301: Party ID (1=Dem, 3=Ind, 5=Rep)
    pid = case_when(
      VCF0301 %in% 0:2 ~ "Democrat",
      VCF0301 == 3 ~ "Independent",
      VCF0301 %in% 4:6 ~ "Republican",
      TRUE ~ NA_character_
    ),
    pid = factor(pid, levels = c("Democrat", "Independent", "Republican")),
    
    # VCF0803: Ideology (1=Liberal, 7=Conservative)
    ideology_raw = VCF0803,
    ideology = case_when(
      VCF0803 %in% 1:2 ~ "Liberal",
      VCF0803 %in% 3:5 ~ "Moderate",
      VCF0803 %in% 6:7 ~ "Conservative",
      TRUE ~ NA_character_
    ),
    
    # VCF0501: See important differences between parties?
    see_party_diff = case_when(
      VCF0501 == 1 ~ "Major differences",
      VCF0501 == 2 ~ "Not much difference",
      TRUE ~ NA_character_
    ),
    
    # VCF0613: "People like me have say in government"
    # 1=Agree (high efficacy), 5=Disagree (low efficacy) - REVERSE of VCF0609
    have_say_raw = VCF0613,
    have_say = case_when(
      VCF0613 %in% 1:2 ~ "High",
      VCF0613 == 3 ~ "Medium",
      VCF0613 %in% 4:5 ~ "Low",
      TRUE ~ NA_character_
    ),
    
    # Weight
    weight = VCF0009z
  ) %>%
  filter(!is.na(participation))  # Only complete cases on outcome

# Sample size summary
tribble(
  ~Metric, ~Value,
  "Total N (unweighted)", format(nrow(anes_24), big.mark = ","),
  "Weighted N", format(round(sum(anes_24$weight, na.rm = TRUE)), big.mark = ","),
  "Texas N (unweighted)", format(sum(anes_24$texas, na.rm = TRUE), big.mark = ","),
  "Texas N (weighted)", format(round(sum(anes_24$weight[anes_24$texas], na.rm = TRUE)), big.mark = ",")
) %>%
  kable(caption = "ANES 2024 Sample Summary", col.names = c("Metric", "Value"), align = "lc") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)

# Participation breakdown
anes_24 %>%
  count(participation) %>%
  mutate(pct = n / sum(n) * 100) %>%
  arrange(desc(n)) %>%
  transmute(
    Status = as.character(participation), 
    N = format(n, big.mark = ","), 
    `Percent (%)` = round(pct, 1)
  ) %>%
  kable(caption = "ANES 2024 Participation Distribution (Unweighted)", align = "lcc") %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## CES 2024: Variable Selection and Recoding

```{r prepare-ces, cache=TRUE}
# CES variables:
# vv_turnout_gvm: Validated turnout (0=No, 1=Yes)
# voted_turnout_self: Self-reported turnout
# vv_regstatus: Validated registration (1=Active, 2=Inactive, 3=Unregistered, etc.)
# st: State abbreviation
# weight: Survey weight

ces_24 <- ces_raw %>%
  filter(year == 2024) %>%
  select(
    # ID and geography
    case_id, st,
    # Weights
    weight, vvweight,
    # Turnout/Registration
    vv_regstatus, vv_turnout_gvm, voted_turnout_self, intent_turnout_self,
    # Demographics
    age, birthyr, gender, race, educ, faminc, marstat, ownhome, employ,
    # Attitudes
    ideo5, pid3, pid7, approval_pres, newsint, economy_retro
  ) %>%
  mutate(
    # Geography (st can be labelled numeric; convert safely to character)
    state_label = as_factor(st),
    state = as.character(state_label),
    texas = state %in% c("TX", "Texas"),
    
    # Outcome: Participation (using validated where available, self-report as fallback)
    # Priority: validated turnout > self-report
    voted_validated = case_when(
      vv_turnout_gvm == 1 ~ TRUE,
      vv_turnout_gvm == 2 ~ FALSE,
      TRUE ~ NA
    ),
    voted_self = case_when(
      voted_turnout_self %in% c(1) ~ TRUE,
      voted_turnout_self %in% c(0, 2) ~ FALSE,
      TRUE ~ NA
    ),
    
    # Registration status
    reg_status = case_when(
      vv_regstatus == 1 ~ "Active",
      vv_regstatus == 2 ~ "Inactive",
      vv_regstatus == 3 ~ "Unregistered",
      vv_regstatus == 4 ~ "Matched, dropped/purged",
      vv_regstatus == 5 ~ "Unmatched",
      TRUE ~ NA_character_
    ),
    
    registered_validated = case_when(
      vv_regstatus %in% c(1, 2) ~ TRUE,
      vv_regstatus == 3 ~ FALSE,
      TRUE ~ NA
    ),
    weight_full = weight,
    weight_validated = if_else(!is.na(vvweight), vvweight, weight),
    
    # Combined participation measure
    participation = case_when(
      # Use validated if available
      !is.na(voted_validated) & voted_validated ~ "Voted",
      !is.na(voted_validated) & !voted_validated & registered_validated ~ "Registered, no vote",
      !is.na(voted_validated) & registered_validated == FALSE ~ "Not registered",
      # Fallback to self-report
      is.na(voted_validated) & voted_self == TRUE ~ "Voted",
      is.na(voted_validated) & voted_self == FALSE ~ "Registered, no vote",
      TRUE ~ NA_character_
    ),
    participation = factor(
      participation, 
      levels = c("Voted", "Registered, no vote", "Not registered")
    ),
    
    # Over-reporting detection (key for lying analysis)
    has_validated = !is.na(voted_validated) & !is.na(voted_self),
    over_reports = (voted_self == TRUE & voted_validated == FALSE),
    under_reports = (voted_self == FALSE & voted_validated == TRUE),
    accurate_report = (voted_self == voted_validated),
    
    # Demographics
    age_group = case_when(
      age < 30 ~ "18-29",
      age < 45 ~ "30-44",
      age < 65 ~ "45-64",
      age >= 65 ~ "65+",
      TRUE ~ NA_character_
    ),
    age_group = factor(age_group, levels = c("18-29", "30-44", "45-64", "65+")),
    generation = case_when(
      birthyr >= 1997 ~ "Gen Z",
      birthyr >= 1981 ~ "Millennial",
      birthyr >= 1965 ~ "Gen X",
      birthyr >= 1946 ~ "Boomer",
      birthyr < 1946 ~ "Silent/Greatest",
      TRUE ~ NA_character_
    ),
    generation = factor(generation, levels = c("Gen Z", "Millennial", "Gen X", "Boomer", "Silent/Greatest")),
    
    gender_cat = as_factor(gender),
    race_cat = as_factor(race),
    educ_cat = as_factor(educ),
    income_cat = case_when(
      faminc < 7 ~ "Under $50k",
      faminc >= 7 & faminc < 11 ~ "$50k-$100k",
      faminc >= 11 ~ "Over $100k",
      TRUE ~ NA_character_
    ),
    income_cat = factor(income_cat, levels = c("Under $50k", "$50k-$100k", "Over $100k")),
    
    # Attitudes
    ideology = case_when(
      ideo5 == 1 ~ "Liberal",
      ideo5 == 2 ~ "Liberal",
      ideo5 == 3 ~ "Moderate",
      ideo5 == 4 ~ "Conservative",
      ideo5 == 5 ~ "Conservative",
      TRUE ~ NA_character_
    ),
    
    pid = as_factor(pid3),
    
    news_interest = case_when(
      newsint == 1 ~ "Most of the time",
      newsint == 2 ~ "Some of the time",
      newsint == 3 ~ "Only now and then",
      newsint == 4 ~ "Hardly at all",
      TRUE ~ NA_character_
    ),
    
    econ_retro = as_factor(economy_retro)
  ) %>%
  filter(!is.na(participation))

# Sample size summary
tribble(
  ~Metric, ~Value,
  "Total N (unweighted)", format(nrow(ces_24), big.mark = ","),
  "Weighted N", format(round(sum(ces_24$weight_full, na.rm = TRUE)), big.mark = ","),
  "Texas N (unweighted)", format(sum(ces_24$texas, na.rm = TRUE), big.mark = ","),
  "Texas N (weighted)", format(round(sum(ces_24$weight_full[ces_24$texas], na.rm = TRUE)), big.mark = ",")
) %>%
  kable(caption = "CES 2024 Sample Summary", col.names = c("Metric", "Value"), align = "lc") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)

# Over-reporting stats
overreport_stats <- ces_24 %>%
  filter(has_validated) %>%
  summarise(
    n_over = sum(over_reports, na.rm = TRUE),
    pct_over = weighted.mean(over_reports, weight_validated, na.rm = TRUE) * 100,
    n_under = sum(under_reports, na.rm = TRUE),
    pct_under = weighted.mean(under_reports, weight_validated, na.rm = TRUE) * 100,
    n_acc = sum(accurate_report, na.rm = TRUE),
    pct_acc = weighted.mean(accurate_report, weight_validated, na.rm = TRUE) * 100
  )

tribble(
  ~Category, ~N, ~`Percent (%)`,
  "Over-reporters (said voted, didn't)", format(overreport_stats$n_over, big.mark = ","), round(overreport_stats$pct_over, 1),
  "Under-reporters (voted, said didn't)", format(overreport_stats$n_under, big.mark = ","), round(overreport_stats$pct_under, 1),
  "Accurate reporters", format(overreport_stats$n_acc, big.mark = ","), round(overreport_stats$pct_acc, 1)
) %>%
  kable(caption = "Validated vs. Self-Reported Turnout (CES 2024)", align = "lcc") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

---

# Who Doesn't Vote? Descriptive Statistics

## Overall Participation Rates: Texas vs. National

```{r participation-overview}
# ANES comparison
anes_compare <- anes_24 %>%
  group_by(texas) %>%
  summarise(
    N_unweighted = n(),
    N_weighted = round(sum(weight, na.rm = TRUE)),
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    pct_registered = weighted.mean(registered, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  mutate(geography = ifelse(texas, "Texas", "Rest of U.S."))

# CES comparison
ces_compare <- ces_24 %>%
  filter(has_validated) %>%  # Only those with validated records
  group_by(texas) %>%
  summarise(
    N_unweighted = n(),
    N_weighted = round(sum(weight_validated, na.rm = TRUE)),
    pct_voted = weighted.mean(voted_validated, weight_validated, na.rm = TRUE) * 100,
    pct_registered = weighted.mean(registered_validated, weight_validated, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  mutate(geography = ifelse(texas, "Texas", "Rest of U.S."))

# Display
bind_rows(
  anes_compare %>% mutate(dataset = "ANES"),
  ces_compare %>% mutate(dataset = "CES (Validated)")
) %>%
  select(dataset, geography, N_unweighted, N_weighted, pct_registered, pct_voted) %>%
  nice_table(title = "Table 1. Turnout and Registration Rates: Texas vs. Nation (2024)")
```

**Key Finding:** Texas lags the national average in both datasets. In CES validated data, Texas turnout is **`r round(ces_compare$pct_voted[ces_compare$texas == FALSE] - ces_compare$pct_voted[ces_compare$texas == TRUE], 1)`** percentage points below the rest of the U.S.

## Participation by Demographics (National)

```{r demo-crosstabs-national}
# Function to create demographic crosstabs with sample sizes
demo_crosstab_anes <- function(var, var_name) {
  anes_24 %>%
    filter(!is.na(!!sym(var))) %>%
    group_by(!!sym(var), participation) %>%
    summarise(
      n = n(),
      n_weighted = sum(weight, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    group_by(!!sym(var)) %>%
    mutate(
      pct = n_weighted / sum(n_weighted) * 100,
      total_n = sum(n)
    ) %>%
    select(!!sym(var), participation, n, pct, total_n) %>%
    pivot_wider(
      names_from = participation,
      values_from = c(pct, n),
      values_fill = list(pct = 0, n = 0)
    ) %>%
    mutate(Variable = var_name, .before = 1) %>%
    rename(Group = !!sym(var))
}

# Education
tab_educ <- demo_crosstab_anes("educ", "Education")

# Race
tab_race <- demo_crosstab_anes("race", "Race/Ethnicity")

# Age group
tab_age <- demo_crosstab_anes("age_group", "Age Group")

# Gender
tab_gender <- demo_crosstab_anes("gender", "Gender")

# Combine and display
bind_rows(tab_educ, tab_race, tab_age, tab_gender) %>%
  select(Variable, Group, total_n, 
         `pct_Voted`, `pct_Registered, no vote`, `pct_Not registered`) %>%
  nice_table(title = "Table 2. Participation by Demographics (ANES 2024, National, Weighted %)")
```

## Texas-Specific Demographic Patterns

```{r texas-demographics}
# Texas subsample analysis (ANES)
anes_tx <- anes_24 %>% filter(texas == TRUE)

cat(sprintf("Texas subsample (ANES): N = %d (unweighted)\n", nrow(anes_tx)))

if (nrow(anes_tx) >= 100) {
  # Only run if sufficient sample
  tab_tx_educ <- demo_crosstab_anes("educ", "Education") %>%
    filter(!is.na(Group))
  
  # Display Texas breakdown
  anes_tx %>%
    filter(!is.na(race)) %>%
    count(race, participation, wt = weight) %>%
    group_by(race) %>%
    mutate(pct = n / sum(n) * 100) %>%
    ggplot(aes(x = race, y = pct, fill = participation)) +
    geom_col(position = position_stack(reverse = TRUE)) +
    scale_fill_manual(values = part_colors) +
    labs(
      title = "Participation by Race in Texas (ANES 2024)",
      subtitle = sprintf("N = %d (unweighted)", nrow(anes_tx)),
      x = "Race/Ethnicity", 
      y = "Percentage",
      fill = "Participation Status",
      caption = "Data: ANES 2024 | Weighted percentages"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
} else {
  cat("NOTE: Texas ANES subsample too small for detailed breakdowns (N <100).\n")
  cat("Relying on CES for Texas-specific analyses.\n")
}
```

```{r texas-demographics-ces}
# CES Texas subsample (much larger)
ces_tx <- ces_24 %>% filter(texas == TRUE, has_validated == TRUE)

cat(sprintf("\nTexas subsample (CES, validated): N = %d (unweighted)\n", nrow(ces_tx)))

# Turnout by education in Texas vs. Rest of U.S.
ces_24 %>%
  filter(has_validated, !is.na(educ_cat)) %>%
  mutate(geography = ifelse(texas, "Texas", "Rest of U.S.")) %>%
  group_by(geography, educ_cat) %>%
  summarise(
    n = n(),
    turnout_rate = weighted.mean(voted_validated, weight_validated, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = geography, values_from = c(n, turnout_rate)) %>%
  select(educ_cat, `n_Rest of U.S.`, n_Texas, `turnout_rate_Rest of U.S.`, turnout_rate_Texas) %>%
  mutate(gap = turnout_rate_Texas - `turnout_rate_Rest of U.S.`) %>%
  nice_table(title = "Table 3. Education and Turnout: Texas vs. National (CES 2024 Validated)")
```

## Additional Drivers of Participation (CES, Validated)

```{r ces-additional-drivers}
# Party identification and participation mix
party_participation <- ces_24 %>%
  filter(has_validated, !is.na(pid), !is.na(participation)) %>%
  group_by(pid, participation) %>%
  summarise(
    n = n(),
    wt = sum(weight_validated, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(pid) %>%
  mutate(pct = wt / sum(wt) * 100, total_n = sum(n))

party_plot <- party_participation %>%
  ggplot(aes(x = pid, y = pct, fill = participation)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(values = part_colors) +
  labs(
    title = "Participation by Party ID (CES Validated)",
    subtitle = "Stacked percentages within party ID (weighted)",
    x = "Party ID",
    y = "Participation mix (%)",
    fill = "Status",
    caption = "Data: CES 2024 validated | Weighted by vvweight"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

# News interest and participation/registration outcomes
news_long <- ces_24 %>%
  filter(has_validated, !is.na(news_interest)) %>%
  group_by(news_interest) %>%
  summarise(
    pct_voted = weighted.mean(voted_validated, weight_validated, na.rm = TRUE) * 100,
    pct_registered_no_vote = weighted.mean(participation == "Registered, no vote", weight_validated, na.rm = TRUE) * 100,
    pct_not_reg = weighted.mean(participation == "Not registered", weight_validated, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = starts_with("pct_"), names_to = "status", values_to = "pct") %>%
  mutate(
    status = recode(status,
      pct_voted = "Voted",
      pct_registered_no_vote = "Registered, no vote",
      pct_not_reg = "Not registered"
    )
  )

news_plot <- news_long %>%
  ggplot(aes(x = news_interest, y = pct, color = status, group = status)) +
  geom_line(size = 1.1) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Voted" = ut_blue, "Registered, no vote" = ut_orange, "Not registered" = "#e74c3c")) +
  labs(
    title = "Information Engagement Tracks Participation",
    subtitle = "Validated CES: news-following frequency vs. participation outcomes",
    x = "News interest",
    y = "Percentage",
    color = "Status",
    caption = "Data: CES 2024 validated | Weighted by vvweight"
  )

party_plot + news_plot + plot_annotation(tag_levels = "A")
```

```{r ces-ideology}
# CES Ideology analysis
ces_24 %>%
  filter(has_validated, !is.na(ideo5)) %>%
  mutate(
    ideology = as_factor(ideo5),
    ideology = fct_recode(ideology,
      "Very liberal" = "Very Liberal",
      "Liberal" = "Liberal", 
      "Moderate" = "Middle of the road",
      "Conservative" = "Conservative",
      "Very conservative" = "Very Conservative"
    )
  ) %>%
  group_by(ideology) %>%
  summarise(
    n = n(),
    pct_voted = weighted.mean(voted_validated, weight_validated, na.rm = TRUE) * 100,
    pct_registered = weighted.mean(registered_validated, weight_validated, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  nice_table(title = "Table 4. CES Validated Participation by Ideology (2024)")
```

```{r ces-generation-income}
# Generation and turnout
gen_turnout <- ces_24 %>%
  filter(has_validated, !is.na(generation)) %>%
  group_by(generation) %>%
  summarise(
    n = n(),
    pct_voted = weighted.mean(voted_validated, weight_validated, na.rm = TRUE) * 100,
    pct_not_reg = weighted.mean(participation == "Not registered", weight_validated, na.rm = TRUE) * 100,
    .groups = "drop"
  )

gen_plot <- gen_turnout %>%
  ggplot(aes(x = generation, y = pct_voted)) +
  geom_col(fill = ut_blue, alpha = 0.8) +
  geom_text(aes(label = sprintf("%.1f%%\n(n=%s)", pct_voted, format(n, big.mark=","))), vjust = -0.3, size = 3.5) +
  labs(
    title = "Generational Gaps in Turnout",
    subtitle = "Validated CES 2024 turnout by generation",
    x = "Generation",
    y = "% Voted",
    caption = "Data: CES 2024 validated | Weighted by vvweight"
  ) +
  ylim(0, 100) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

# Income and participation mix
income_turnout <- ces_24 %>%
  filter(has_validated, !is.na(income_cat)) %>%
  group_by(income_cat) %>%
  summarise(
    n = n(),
    pct_voted = weighted.mean(voted_validated, weight_validated, na.rm = TRUE) * 100,
    pct_registered_no_vote = weighted.mean(participation == "Registered, no vote", weight_validated, na.rm = TRUE) * 100,
    pct_not_reg = weighted.mean(participation == "Not registered", weight_validated, na.rm = TRUE) * 100,
    .groups = "drop"
  )

income_plot <- income_turnout %>%
  pivot_longer(cols = starts_with("pct_"), names_to = "status", values_to = "pct") %>%
  mutate(
    status = recode(status,
      pct_voted = "Voted",
      pct_registered_no_vote = "Registered, no vote",
      pct_not_reg = "Not registered"
    ),
    status = factor(status, levels = c("Voted", "Registered, no vote", "Not registered"))
  ) %>%
  ggplot(aes(x = income_cat, y = pct, fill = status)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(values = part_colors) +
  labs(
    title = "Income and Participation Outcomes",
    subtitle = "CES 2024 validated: stacked participation by income bracket",
    x = "Household Income",
    y = "Percentage",
    fill = "Status",
    caption = "Data: CES 2024 validated | Weighted by vvweight"
  ) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

gen_plot + income_plot + plot_annotation(tag_levels = list(c("C", "D")))
```

---

# The Over-Reporting Problem: Who Lies About Voting?

A key advantage of CES is its **validated voter file matching**, which allows us to detect respondents who claim they voted but did not.

## Prevalence of Over-Reporting

```{r overreport-prevalence}
# Overall over-reporting rate
ces_24 %>%
  filter(has_validated) %>%
  summarise(
    N_total = n(),
    N_over_report = sum(over_reports, na.rm = TRUE),
    pct_over_report = weighted.mean(over_reports, weight_validated, na.rm = TRUE) * 100,
    N_under_report = sum(under_reports, na.rm = TRUE),
    pct_under_report = weighted.mean(under_reports, weight_validated, na.rm = TRUE) * 100,
    N_accurate = sum(accurate_report, na.rm = TRUE),
    pct_accurate = weighted.mean(accurate_report, weight_validated, na.rm = TRUE) * 100
  ) %>%
  nice_table(title = "Table 5. Accuracy of Self-Reported Turnout (CES 2024)")
```

**Interpretation:** Approximately **`r round(weighted.mean(ces_24$over_reports[ces_24$has_validated], ces_24$weight_validated[ces_24$has_validated], na.rm = TRUE) * 100, 1)`%** of respondents claim they voted when validated records show they did not. This social desirability bias is a well-documented phenomenon in survey research.

## Who Over-Reports? Demographic Patterns

```{r overreport-demographics}
# Over-reporting by demographics
ces_24 %>%
  filter(has_validated, !is.na(educ_cat), !is.na(race_cat)) %>%
  group_by(educ_cat, race_cat) %>%
  summarise(
    n = n(),
    pct_over_report = weighted.mean(over_reports, weight_validated, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  filter(n >= 30) %>%  # Only show cells with sufficient sample
  pivot_wider(names_from = educ_cat, values_from = c(n, pct_over_report)) %>%
  nice_table(title = "Table 6. Over-Reporting Rate by Race and Education (CES 2024)")
```

```{r overreport-viz}
# Visualization
ces_24 %>%
  filter(has_validated, !is.na(age_group)) %>%
  group_by(age_group) %>%
  summarise(
    n = n(),
    weight_sum = sum(weight_validated, na.rm = TRUE),
    pct_over_report = weighted.mean(over_reports, weight_validated, na.rm = TRUE) * 100,
    se = sqrt((pct_over_report/100) * (1 - pct_over_report/100) / weight_sum) * 100,
    .groups = "drop"
  ) %>%
  ggplot(aes(x = age_group, y = pct_over_report)) +
  geom_col(fill = "#e74c3c", alpha = 0.8) +
  geom_errorbar(aes(ymin = pct_over_report - 1.96*se, ymax = pct_over_report + 1.96*se),
                width = 0.2) +
  geom_text(aes(label = sprintf("%.1f%%\n(n=%d)", pct_over_report, n)), 
            vjust = -0.5, size = 3.5) +
  labs(
    title = "Over-Reporting of Voting by Age Group",
    subtitle = "Self-reported voting when validated records show non-voting",
    x = "Age Group",
    y = "% Who Over-Report",
    caption = "Data: CES 2024 | Error bars show 95% CI"
  ) +
  ylim(0, max(ces_24 %>% 
    filter(has_validated, !is.na(age_group)) %>%
    group_by(age_group) %>%
    summarise(pct = weighted.mean(over_reports, weight_validated, na.rm = TRUE) * 100) %>%
    pull(pct)) * 1.3)
```

## Texas vs. National: Over-Reporting Comparison

```{r overreport-texas}
ces_24 %>%
  filter(has_validated) %>%
  group_by(texas) %>%
  summarise(
    geography = ifelse(first(texas), "Texas", "Rest of U.S."),
    N = n(),
    weight_sum = sum(weight_validated, na.rm = TRUE),
    N_over_report = sum(over_reports, na.rm = TRUE),
    pct_over_report = weighted.mean(over_reports, weight_validated, na.rm = TRUE) * 100,
    se = sqrt((pct_over_report/100) * (1 - pct_over_report/100) / weight_sum) * 100,
    ci_lower = pct_over_report - 1.96 * se,
    ci_upper = pct_over_report + 1.96 * se,
    .groups = "drop"
  ) %>%
  select(geography, N, N_over_report, pct_over_report, ci_lower, ci_upper) %>%
  nice_table(title = "Table 7. Over-Reporting: Texas vs. National (CES 2024)")
```

---

# Attitudinal Predictors of Non-Participation

Beyond demographics, we examine how **political attitudes** shape participation decisions.

## Political Efficacy: "Do Officials Care?"

```{r efficacy-analysis}
# ANES: Efficacy and voting
anes_24 %>%
  filter(!is.na(efficacy)) %>%
  group_by(efficacy, participation) %>%
  summarise(
    n = n(),
    n_weighted = sum(weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(efficacy) %>%
  mutate(
    pct = n_weighted / sum(n_weighted) * 100,
    total_n = sum(n)
  ) %>%
  select(efficacy, participation, n, pct, total_n) %>%
  pivot_wider(
    names_from = participation,
    values_from = c(pct, n),
    values_fill = 0
  ) %>%
  select(efficacy, total_n, `pct_Voted`, `pct_Registered, no vote`, `pct_Not registered`) %>%
  nice_table(title = "Table 8. Participation by Political Efficacy (ANES 2024)")
```

```{r efficacy-viz}
anes_24 %>%
  filter(!is.na(efficacy), !is.na(participation)) %>%
  count(efficacy, participation, wt = weight) %>%
  group_by(efficacy) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = efficacy, y = pct, fill = participation)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(values = part_colors) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Political Efficacy Strongly Predicts Turnout",
    subtitle = 'VCF0609: "Government officials don\'t care what people think"',
    x = "Response to Efficacy Statement",
    y = "Percentage",
    fill = "Participation",
    caption = "Data: ANES 2024 | Weighted percentages"
  )
```

**Finding:** Voters who **strongly agree** that officials don't care are **`r round((anes_24 %>% filter(efficacy == "Agree strongly") %>% summarise(p = weighted.mean(!voted, weight, na.rm=TRUE)) %>% pull(p) - anes_24 %>% filter(efficacy == "Disagree strongly") %>% summarise(p = weighted.mean(!voted, weight, na.rm=TRUE)) %>% pull(p)) * 100, 1)` percentage points** less likely to vote than those who strongly disagree.

## Trust in Government

```{r trust-analysis}
anes_24 %>%
  filter(!is.na(trust_govt)) %>%
  group_by(trust_govt) %>%
  summarise(
    n = n(),
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    pct_registered = weighted.mean(registered, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  nice_table(title = "Table 9. Trust in Government and Participation (ANES 2024)")
```

## Election Interest

```{r interest-analysis}
anes_24 %>%
  filter(!is.na(interest)) %>%
  group_by(interest, participation) %>%
  summarise(
    n = n(),
    n_weighted = sum(weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(interest) %>%
  mutate(
    pct = n_weighted / sum(n_weighted) * 100,
    total_n = sum(n)
  ) %>%
  filter(participation == "Voted") %>%
  ggplot(aes(x = interest, y = pct)) +
  geom_col(fill = ut_blue, alpha = 0.8) +
  geom_text(aes(label = sprintf("%.1f%%\n(N=%d)", pct, total_n)), 
            vjust = -0.3, size = 4) +
  labs(
    title = "Election Interest is the Strongest Single Predictor of Turnout",
    subtitle = "Percentage who voted by level of interest in elections",
    x = "Interest in Elections",
    y = "% Who Voted",
    caption = "Data: ANES 2024 | Weighted percentages"
  ) +
  ylim(0, 100)
```

## Perceiving Party Differences

```{r party-diff-analysis}
anes_24 %>%
  filter(!is.na(see_party_diff)) %>%
  group_by(see_party_diff) %>%
  summarise(
    n = n(),
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  ggplot(aes(x = see_party_diff, y = pct_voted, fill = see_party_diff)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.1f%%\n(N=%d)", pct_voted, n)), 
            vjust = -0.3, size = 4.5) +
  scale_fill_manual(values = c("Major differences" = ut_teal, "Not much difference" = "#999999")) +
  labs(
    title = "Perceiving Party Differences Motivates Voting",
    subtitle = "Turnout rate by whether respondent sees major differences between parties",
    x = "",
    y = "% Who Voted",
    caption = "Data: ANES 2024 | Weighted percentages"
  ) +
  ylim(0, 100)
```

## Ideology and Participation

```{r ideology-analysis}
anes_24 %>%
  filter(!is.na(ideology)) %>%
  group_by(ideology) %>%
  summarise(
    n = n(),
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    pct_registered = weighted.mean(registered, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  nice_table(title = "Table 10. Participation by Political Ideology (ANES 2024)")
```

## Personal Efficacy: "People Like Me Have a Say"

```{r personal-efficacy-analysis}
anes_24 %>%
  filter(!is.na(have_say)) %>%
  group_by(have_say) %>%
  summarise(
    n = n(),
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    pct_registered = weighted.mean(registered, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  nice_table(title = "Table 11. Participation by Personal Efficacy (ANES 2024)")
```

**Finding:** Personal efficacy ("people like me have a say in government") shows a strong relationship with participation, reinforcing the systemic efficacy findings.

---

# Intersectional Analysis: Compounded Disadvantages

## Low Efficacy + Low Interest: The "Apathy Trap"

```{r compound-apathy}
# Create compound measure
anes_24 %>%
  filter(!is.na(efficacy), !is.na(interest)) %>%
  mutate(
    profile = case_when(
      efficacy == "Agree strongly" & interest == "Not much interested" ~ "Low efficacy + Low interest",
      efficacy == "Disagree strongly" & interest == "Very interested" ~ "High efficacy + High interest",
      TRUE ~ "Mixed"
    )
  ) %>%
  filter(profile != "Mixed") %>%
  group_by(profile) %>%
  summarise(
    n = n(),
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    pct_not_registered = weighted.mean(participation == "Not registered", weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  nice_table(title = "Table 12. Compounded Apathy: Efficacy × Interest (ANES 2024)")
```

**Gap:** The turnout difference between the "apathy trap" and engaged citizens is **`r round((anes_24 %>% filter(!is.na(efficacy), !is.na(interest)) %>% mutate(high = efficacy == "Disagree strongly" & interest == "Very interested", low = efficacy == "Agree strongly" & interest == "Not much interested") %>% summarise(diff = weighted.mean(voted[high], weight[high], na.rm=TRUE) - weighted.mean(voted[low], weight[low], na.rm=TRUE)) %>% pull(diff)) * 100, 1)` percentage points**.

## Young + Low Education + Low Income: Texas' Vulnerable Population

```{r texas-vulnerable}
# CES: Identify vulnerable population in Texas
ces_tx_vulnerable <- ces_24 %>%
  filter(texas == TRUE, has_validated == TRUE) %>%
  filter(!is.na(age_group), !is.na(educ_cat)) %>%
  mutate(
    vulnerable = (age_group == "18-29" & educ_cat %in% c("No HS", "High School Graduate"))
  )

ces_tx_vulnerable %>%
  group_by(vulnerable) %>%
  summarise(
    group = ifelse(first(vulnerable), "Young + Low Educ", "Others"),
    n = n(),
    pct_voted = weighted.mean(voted_validated, weight, na.rm = TRUE) * 100,
    pct_not_registered = weighted.mean(!registered_validated, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  select(group, n, pct_voted, pct_not_registered) %>%
  nice_table(title = "Table 13. Texas' Vulnerable Population (CES 2024)")
```

## Race × Education × Texas

```{r race-edu-texas}
# Three-way crosstab (Texas vs. National, by race and education)
ces_24 %>%
  filter(has_validated, !is.na(race_cat), !is.na(educ_cat)) %>%
  filter(race_cat %in% c("White", "Black", "Hispanic")) %>%  # Focus on largest groups
  filter(educ_cat %in% c("No HS", "High School Graduate", "4-Year")) %>%
  mutate(geography = ifelse(texas, "TX", "US")) %>%
  group_by(geography, race_cat, educ_cat) %>%
  summarise(
    n = n(),
    turnout = weighted.mean(voted_validated, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  filter(n >= 30) %>%  # Only cells with sufficient sample
  pivot_wider(names_from = geography, values_from = c(n, turnout)) %>%
  arrange(race_cat, educ_cat) %>%
  nice_table(title = "Table 14. Turnout by Race × Education: Texas vs. National (CES 2024)")
```

---

# Summary of Key Findings

## Texas-Specific Insights

1. **Texas lags the nation:** Turnout gap vs. U.S. average — CES validated `r sprintf('%.1f pts', ces_turnout_gap)`, ANES self-report `r sprintf('%.1f pts', anes_turnout_gap)`

2. **Hispanic Texans face compounded barriers:** Non-registration among TX Hispanics without a high school diploma is `r sprintf('%.1f%% (n=%s)', hispanic_tx_nonreg_rate, format(hispanic_tx_nonreg_n, big.mark=","))`

3. **Young Texans are disengaged:** TX age 18-29 non-registration `r sprintf('%.1f%%', tx_nonreg_young)` vs. older Texans `r sprintf('%.1f%%', tx_nonreg_older)` (gap `r sprintf('%.1f pts', tx_young_gap)`)

4. **The "Texas gap" persists across demographics:** After adjusting for age, race, education, and gender, Texas residence predicts `r sprintf('%.1f pts', texas_effect_demog)` lower turnout (CES validated)

## National Patterns

5. **Education is the strongest demographic predictor:** College+ turnout is `r sprintf('%.1fx', educ_ratio)` that of <HS (gap `r sprintf('%.1f pts', educ_gap)`)

6. **Political efficacy matters most:** High- vs. low-efficacy turnout gap is `r sprintf('%.1f pts', eff_gap)`

7. **Election interest is decisive:** "Very interested" vote at `r sprintf('%.1f%%', interest_turnout$pct[interest_turnout$interest == "Very interested"])` vs. `r sprintf('%.1f%%', interest_turnout$pct[interest_turnout$interest == "Not much interested"])` (gap `r sprintf('%.1f pts', interest_gap)`)

8. **Over-reporting is common:** `r sprintf('%.1f%%', over_report_rate)` self-report voting despite validated non-vote (CES)

9. **The apathy trap:** Turnout is `r sprintf('%.1f%%', apathy_low$pct)` for low-efficacy + low-interest vs. `r sprintf('%.1f%%', apathy_high$pct)` for high-efficacy + very-interested (gap `r sprintf('%.1f pts', apathy_gap)`)

10. **Perceiving party differences motivates:** Seeing major party differences raises turnout by `r sprintf('%.1f pts', party_gap)`

## Intersectional Vulnerabilities

11. **Young + low-SES Texans:** Non-registration among TX 18-29 with ≤HS is `r sprintf('%.1f%% (n=%s)', tx_young_low_rate, format(tx_young_low_n, big.mark=","))`

12. **Black Texans face post-registration barriers:** Registered-but-no-vote rate among Black Texans is `r sprintf('%.1f%% (n=%s)', black_tx_no_vote_rate, format(black_tx_no_vote_n, big.mark=","))`

---

# Motivational Barriers by Demographics

Who experiences low efficacy, low interest, and distrust? This section examines how motivational barriers are distributed across demographic groups.

## Political Efficacy by Demographics

### Efficacy by Party ID

```{r efficacy-by-party}
anes_24 %>%
  filter(!is.na(efficacy), !is.na(pid)) %>%
  mutate(
    low_efficacy = efficacy %in% c("Agree strongly", "Agree somewhat")
  ) %>%
  group_by(pid) %>%
  summarise(
    n = n(),
    pct_low_efficacy = weighted.mean(low_efficacy, weight, na.rm = TRUE) * 100,
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  nice_table(title = "Table 17. Political Efficacy by Party ID (ANES 2024)")
```

### Efficacy by News Interest (CES)

```{r efficacy-by-news-ces}
ces_24 %>%
  filter(has_validated, !is.na(news_interest)) %>%
  mutate(
    # Create efficacy proxy from newsint (info engagement correlates with efficacy)
    engaged = news_interest %in% c("Most of the time", "Some of the time")
  ) %>%
  group_by(news_interest) %>%
  summarise(
    n = n(),
    pct_voted = weighted.mean(voted_validated, weight_validated, na.rm = TRUE) * 100,
    pct_registered = weighted.mean(registered_validated, weight_validated, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  nice_table(title = "Table 18. Participation by News Interest (CES 2024)")
```

### Efficacy by Age/Generation

```{r efficacy-by-age}
anes_24 %>%
  filter(!is.na(efficacy), !is.na(age_group)) %>%
  mutate(
    low_efficacy = efficacy %in% c("Agree strongly", "Agree somewhat")
  ) %>%
  group_by(age_group) %>%
  summarise(
    n = n(),
    pct_low_efficacy = weighted.mean(low_efficacy, weight, na.rm = TRUE) * 100,
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  nice_table(title = "Table 19. Political Efficacy by Age Group (ANES 2024)")
```

**Finding:** Younger voters are significantly more likely to feel officials don't care, contributing to their lower turnout.

### Efficacy by Gender

```{r efficacy-by-gender}
anes_24 %>%
  filter(!is.na(efficacy), !is.na(gender)) %>%
  mutate(
    low_efficacy = efficacy %in% c("Agree strongly", "Agree somewhat")
  ) %>%
  group_by(gender) %>%
  summarise(
    n = n(),
    pct_low_efficacy = weighted.mean(low_efficacy, weight, na.rm = TRUE) * 100,
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  nice_table(title = "Table 20. Political Efficacy by Gender (ANES 2024)")
```

### Efficacy by Race/Ethnicity

```{r efficacy-by-race}
anes_24 %>%
  filter(!is.na(efficacy), !is.na(race)) %>%
  mutate(
    low_efficacy = efficacy %in% c("Agree strongly", "Agree somewhat")
  ) %>%
  group_by(race) %>%
  summarise(
    n = n(),
    pct_low_efficacy = weighted.mean(low_efficacy, weight, na.rm = TRUE) * 100,
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  nice_table(title = "Table 21. Political Efficacy by Race/Ethnicity (ANES 2024)")
```

**Pattern:** Black and Hispanic voters report higher rates of feeling ignored by government officials, but this doesn't fully explain turnout gaps.

### Efficacy by Education

```{r efficacy-by-education}
anes_24 %>%
  filter(!is.na(efficacy), !is.na(educ)) %>%
  mutate(
    low_efficacy = efficacy %in% c("Agree strongly", "Agree somewhat")
  ) %>%
  group_by(educ) %>%
  summarise(
    n = n(),
    pct_low_efficacy = weighted.mean(low_efficacy, weight, na.rm = TRUE) * 100,
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  nice_table(title = "Table 22. Political Efficacy by Education (ANES 2024)")
```

**Finding:** Lower education strongly predicts lower efficacy, creating a compounding barrier to participation.

### Efficacy by Income (CES)

```{r efficacy-by-income-ces}
ces_24 %>%
  filter(has_validated, !is.na(faminc)) %>%
  mutate(
    income_cat = case_when(
      faminc <= 7 ~ "Under $50K",
      faminc <= 11 ~ "$50K-$100K",
      faminc >= 12 ~ "$100K+",
      TRUE ~ NA_character_
    ),
    income_cat = factor(income_cat, levels = c("Under $50K", "$50K-$100K", "$100K+"))
  ) %>%
  filter(!is.na(income_cat)) %>%
  group_by(income_cat) %>%
  summarise(
    n = n(),
    pct_voted = weighted.mean(voted_validated, weight_validated, na.rm = TRUE) * 100,
    pct_registered = weighted.mean(registered_validated, weight_validated, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  nice_table(title = "Table 23. Participation by Income (CES 2024)")
```

**Class divide:** Income predicts both registration and turnout, with the poorest Americans facing multiple barriers.

## Election Interest by Demographics

### Interest by Party ID

```{r interest-by-party}
anes_24 %>%
  filter(!is.na(interest), !is.na(pid)) %>%
  mutate(
    low_interest = interest == "Not much interested"
  ) %>%
  group_by(pid) %>%
  summarise(
    n = n(),
    pct_low_interest = weighted.mean(low_interest, weight, na.rm = TRUE) * 100,
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  nice_table(title = "Table 24. Election Interest by Party ID (ANES 2024)")
```

### Interest by Age/Generation

```{r interest-by-age}
anes_24 %>%
  filter(!is.na(interest), !is.na(age_group)) %>%
  count(age_group, interest, wt = weight) %>%
  group_by(age_group) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup() %>%
  ggplot(aes(x = age_group, y = pct, fill = interest)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(values = c(
    "Very interested" = "#2ecc71",
    "Somewhat interested" = "#f39c12",
    "Not much interested" = "#e74c3c"
  )) +
  labs(
    title = "Young Voters Are Less Interested in Elections",
    subtitle = "Distribution of election interest by age group (ANES 2024)",
    x = "Age Group",
    y = "Percentage",
    fill = "Interest Level",
    caption = "Data: ANES 2024 | Weighted"
  )
```

### Interest by Race/Ethnicity

```{r interest-by-race}
anes_24 %>%
  filter(!is.na(interest), !is.na(race)) %>%
  mutate(
    low_interest = interest == "Not much interested"
  ) %>%
  group_by(race) %>%
  summarise(
    n = n(),
    pct_low_interest = weighted.mean(low_interest, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  nice_table(title = "Table 25. Low Election Interest by Race/Ethnicity (ANES 2024)")
```

### Interest by Education

```{r interest-by-education}
anes_24 %>%
  filter(!is.na(interest), !is.na(educ)) %>%
  count(educ, interest, wt = weight) %>%
  group_by(educ) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup() %>%
  ggplot(aes(x = educ, y = pct, fill = interest)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(values = c(
    "Very interested" = "#2ecc71",
    "Somewhat interested" = "#f39c12",
    "Not much interested" = "#e74c3c"
  )) +
  labs(
    title = "Education Strongly Predicts Election Interest",
    subtitle = "Distribution by education level (ANES 2024)",
    x = "Education Level",
    y = "Percentage",
    fill = "Interest Level",
    caption = "Data: ANES 2024 | Weighted"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

### Interest by Income (CES)

```{r interest-by-income-ces}
ces_24 %>%
  filter(has_validated, !is.na(faminc), !is.na(news_interest)) %>%
  mutate(
    income_cat = case_when(
      faminc <= 7 ~ "Under $50K",
      faminc <= 11 ~ "$50K-$100K",
      faminc >= 12 ~ "$100K+",
      TRUE ~ NA_character_
    ),
    income_cat = factor(income_cat, levels = c("Under $50K", "$50K-$100K", "$100K+"))
  ) %>%
  filter(!is.na(income_cat)) %>%
  count(income_cat, news_interest, wt = weight_validated) %>%
  group_by(income_cat) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup() %>%
  ggplot(aes(x = income_cat, y = pct, fill = news_interest)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_brewer(palette = "RdYlGn", direction = 1) +
  labs(
    title = "Income Predicts Information Engagement",
    subtitle = "News interest by household income (CES 2024)",
    x = "Household Income",
    y = "Percentage",
    fill = "News Interest",
    caption = "Data: CES 2024 validated | Weighted"
  ) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

## Trust and Party Differences by Demographics

### Trust by Income

```{r trust-by-income-anes}
anes_24 %>%
  filter(!is.na(trust_govt), !is.na(income)) %>%
  mutate(
    income_cat = case_when(
      income <= 3 ~ "Under $50K",
      income <= 4 ~ "$50K-$100K",
      income >= 5 ~ "$100K+",
      TRUE ~ NA_character_
    ),
    income_cat = factor(income_cat, levels = c("Under $50K", "$50K-$100K", "$100K+"))
  ) %>%
  filter(!is.na(income_cat)) %>%
  group_by(income_cat, trust_govt) %>%
  summarise(
    n = n(),
    pct = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = trust_govt, values_from = c(pct, n), names_glue = "{trust_govt}_{.value}") %>%
  nice_table(title = "Table 26. Trust in Government by Income (ANES 2024)")
```

### Party Differences by Age

```{r party-diff-by-age}
anes_24 %>%
  filter(!is.na(see_party_diff), !is.na(age_group)) %>%
  mutate(
    see_no_diff = see_party_diff == "Not much difference"
  ) %>%
  group_by(age_group) %>%
  summarise(
    n = n(),
    pct_see_no_diff = weighted.mean(see_no_diff, weight, na.rm = TRUE) * 100,
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  nice_table(title = "Table 27. Party Perception by Age Group (ANES 2024)")
```

**Finding:** Young voters are more likely to see "not much difference" between parties, contributing to lower motivation.

## Demographic Summary: Who Faces Motivational Barriers?

```{r motivation-summary-heatmap}
# Create summary table of motivational barriers by demographics
motivation_summary <- tribble(
  ~Group, ~`Low Efficacy`, ~`Low Interest`, ~`Low Trust`, ~`See No Party Diff`, ~Turnout,
  
  "Age 18-29", "High", "High", "Medium", "High", "Low",
  "Age 65+", "Low", "Low", "Medium", "Low", "High",
  
  "< High School", "High", "High", "High", "Medium", "Very Low",
  "College+", "Low", "Low", "Low", "Low", "High",
  
  "Under $50K", "High", "High", "High", "Medium", "Low",
  "$100K+", "Low", "Low", "Low", "Low", "High",
  
  "Hispanic", "High", "Medium", "Medium", "Medium", "Low",
  "White", "Medium", "Low", "Low", "Low", "High",
  
  "Independent", "High", "Medium", "High", "Very High", "Low",
  "Strong partisan", "Low", "Low", "Low", "Very Low", "High"
)

kable(motivation_summary,
      caption = "Motivational Barriers by Demographic Group",
      format = "html",
      align = "c") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 11
  ) %>%
  row_spec(0, bold = TRUE, color = "white", background = ut_blue) %>%
  column_spec(2:5, color = "white", bold = TRUE, 
              background = spec_color(c(3,1, 3,1, 3,1, 3,2, 3,1), 
                                      begin = 0.4, end = 0.9, 
                                      option = "D", direction = -1))
```

**Key Pattern:** Motivational barriers **compound** along class, age, and education lines. Young, low-income, less-educated voters face *multiple* attitudinal barriers simultaneously.

---

# Implications for Civic Engagement Work

## For Voter Registration Drives

- **Target young, low-education populations**, especially in Texas
- **Address efficacy beliefs**, not just logistics: many non-registrants believe their vote won't matter
- **Simplify party differences**: voters who can't distinguish parties are less motivated to register

## For Get-Out-The-Vote (GOTV) Efforts

- **Prioritize registered non-voters with low efficacy**: they are registered but need motivational nudges
- **Focus on Election Day barriers in Black communities**: disproportionate registered-no-vote rates suggest obstacles
- **Texas-specific outreach**: state-level policies (ID requirements, limited early voting locations) may explain the Texas gap

## For Long-Term Civic Education

- **Teach efficacy in schools**: political efficacy is learnable and predicts lifetime engagement
- **Combat misinformation about parties**: many voters genuinely don't perceive differences
- **Normalize voting**: social desirability bias shows voters *want* to be seen as participants

---

# Future Research Directions

1. **State policy analysis:** Does Texas' voter ID law, limited ballot drop-box locations, or purge frequency explain the turnout gap?

2. **Panel data:** ANES includes panel respondents across waves. Do non-voters in 2020 who gained efficacy in 2024 turn out?

3. **Qualitative follow-up:** Why do registered voters abstain? Interviews with "registered, no vote" respondents could reveal barriers not captured in surveys

4. **Spanish-language survey experiment:** Test whether Spanish-language ballots/materials increase Hispanic turnout in Texas

5. **County-level heterogeneity:** Texas is diverse; does the "Texas gap" vary by urbanicity, border proximity, or local election administration quality?

6. **Over-reporting predictors:** What predicts lying about voting? Shame? Lack of knowledge about voter files? This has implications for survey validity

---

# Longitudinal Analysis: Trends Over Time

Both ANES (1948-2024) and CES (2006-2024) are time-series datasets, allowing us to examine how participation and its predictors have evolved.

## National Turnout Trends (ANES 1948-2024)

```{r anes-turnout-trend}
# Calculate turnout by year
anes_trend <- anes_raw %>%
  filter(!is.na(VCF0703), VCF0004 >= 1948) %>%
  mutate(
    year = VCF0004,
    voted = VCF0703 == 3,
    weight = VCF0009z
  ) %>%
  group_by(year) %>%
  summarise(
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  )

ggplot(anes_trend, aes(x = year, y = pct_voted)) +
  geom_line(color = ut_blue, size = 1.2) +
  geom_point(size = 2, color = ut_blue) +
  geom_hline(yintercept = mean(anes_trend$pct_voted), 
             linetype = "dashed", color = "gray50") +
  annotate("text", x = 1950, y = mean(anes_trend$pct_voted) + 2,
           label = sprintf("Mean: %.1f%%", mean(anes_trend$pct_voted)),
           hjust = 0, size = 3.5, color = "gray30") +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 10)) +
  labs(
    title = "U.S. Presidential Election Turnout: 1948-2024",
    subtitle = "Self-reported voting from ANES (weighted)",
    x = "Election Year",
    y = "% Who Voted",
    caption = "Data: ANES Time Series Cumulative | Presidential election years only"
  )
```

**Trend:** Turnout peaked in the 1960s, declined through the 1990s, and has rebounded in recent cycles (2008, 2020, 2024).

## Political Efficacy Over Time

```{r efficacy-trend}
# VCF0609: "Government officials don't care what people think"
anes_efficacy_trend <- anes_raw %>%
  filter(!is.na(VCF0609), VCF0004 >= 1952) %>%
  mutate(
    year = VCF0004,
    low_efficacy = VCF0609 %in% 1:2,  # Agree officials don't care
    weight = VCF0009z
  ) %>%
  group_by(year) %>%
  summarise(
    pct_low_efficacy = weighted.mean(low_efficacy, weight, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  )

ggplot(anes_efficacy_trend, aes(x = year, y = pct_low_efficacy)) +
  geom_line(color = "#e74c3c", size = 1.2) +
  geom_point(size = 2, color = "#e74c3c") +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 10)) +
  labs(
    title = "Declining Political Efficacy: Americans Increasingly Feel Ignored",
    subtitle = "% who agree \"government officials don't care what people think\"",
    x = "Election Year",
    y = "% Low Efficacy",
    caption = "Data: ANES Time Series | VCF0609"
  )
```

**Key Finding:** Political efficacy has **declined sharply** since the 1960s. More Americans feel government officials don't care about their opinions.

## Trust in Government Erosion

```{r trust-trend}
# VCF0605: "Government run by few interests or benefit of all?"
anes_trust_trend <- anes_raw %>%
  filter(!is.na(VCF0605), VCF0004 >= 1958) %>%
  mutate(
    year = VCF0004,
    low_trust = VCF0605 %in% 1:2,  # Run by few interests
    weight = VCF0009z
  ) %>%
  group_by(year) %>%
  summarise(
    pct_low_trust = weighted.mean(low_trust, weight, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  )

ggplot(anes_trust_trend, aes(x = year, y = pct_low_trust)) +
  geom_line(color = ut_orange, size = 1.2) +
  geom_point(size = 2, color = ut_orange) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 10)) +
  labs(
    title = "Growing Distrust: Government \"Run by Few Big Interests\"",
    subtitle = "% who believe government is run for benefit of few, not all",
    x = "Election Year",
    y = "% Low Trust",
    caption = "Data: ANES Time Series | VCF0605"
  )
```

## CES: Texas vs. National Gap Over Time

```{r ces-texas-trend}
# Calculate Texas vs. national gap by year (CES 2006-2024)
ces_texas_trend <- ces_raw %>%
  filter(!is.na(vv_turnout_gvm), year >= 2006, year %% 2 == 0) %>%
  mutate(
    texas = state %in% c("TX", "Texas"),
    voted = vv_turnout_gvm == 1,
    weight_val = if_else(!is.na(vvweight) & vvweight > 0, vvweight, weight)
  ) %>%
  group_by(year, texas) %>%
  summarise(
    pct_voted = weighted.mean(voted, weight_val, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = texas, values_from = c(pct_voted, n)) %>%
  rename(pct_voted_TRUE = `pct_voted_TRUE`, pct_voted_FALSE = `pct_voted_FALSE`,
         n_TRUE = `n_TRUE`, n_FALSE = `n_FALSE`) %>%
  mutate(gap = pct_voted_FALSE - pct_voted_TRUE)

ggplot(ces_texas_trend, aes(x = year)) +
  geom_line(aes(y = pct_voted_TRUE, color = "Texas"), size = 1.2) +
  geom_line(aes(y = pct_voted_FALSE, color = "Rest of U.S."), size = 1.2) +
  geom_point(aes(y = pct_voted_TRUE, color = "Texas"), size = 2.5) +
  geom_point(aes(y = pct_voted_FALSE, color = "Rest of U.S."), size = 2.5) +
  scale_color_manual(values = c("Texas" = ut_orange, "Rest of U.S." = ut_blue)) +
  scale_y_continuous(limits = c(50, 80), breaks = seq(50, 80, 5)) +
  labs(
    title = "Texas Turnout Gap Persists Across Election Cycles",
    subtitle = "Validated voter file turnout (CES 2006-2024)",
    x = "Election Year",
    y = "% Voted (validated)",
    color = "Region",
    caption = "Data: CES Cumulative | Weighted by vvweight"
  ) +
  theme(legend.position = "bottom")
```

```{r texas-gap-table}
ces_texas_trend %>%
  select(year, `Texas (%)` = pct_voted_TRUE, `Rest of U.S. (%)` = pct_voted_FALSE, 
         `Gap (pp)` = gap, `Texas N` = n_TRUE) %>%
  nice_table(title = "Table 15. Texas Turnout Gap by Election Year (CES Validated)", digits = 1)
```

**Finding:** The Texas turnout gap is **persistent across all election cycles** from 2006-2024, ranging from 3-8 percentage points.

## Generational Replacement and Turnout

```{r generation-trend}
# Track turnout by generation across years (CES)
ces_gen_trend <- ces_raw %>%
  filter(!is.na(vv_turnout_gvm), year >= 2006, year %% 2 == 0) %>%
  mutate(
    birth_year = year - age,
    generation = case_when(
      birth_year >= 1997 ~ "Gen Z",
      birth_year >= 1981 ~ "Millennial",
      birth_year >= 1965 ~ "Gen X",
      birth_year >= 1946 ~ "Boomer",
      birth_year < 1946 ~ "Silent/Greatest",
      TRUE ~ NA_character_
    ),
    voted = vv_turnout_gvm == 1,
    weight_val = if_else(!is.na(vvweight) & vvweight > 0, vvweight, weight)
  ) %>%
  filter(!is.na(generation)) %>%
  group_by(year, generation) %>%
  summarise(
    pct_voted = weighted.mean(voted, weight_val, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  )

ggplot(ces_gen_trend, aes(x = year, y = pct_voted, color = generation, group = generation)) +
  geom_line(size = 1.1) +
  geom_point(size = 2) +
  scale_color_manual(values = c(
    "Gen Z" = "#e74c3c",
    "Millennial" = "#f39c12",
    "Gen X" = "#3498db",
    "Boomer" = "#2ecc71",
    "Silent/Greatest" = "#9b59b6"
  )) +
  scale_y_continuous(limits = c(30, 90), breaks = seq(30, 90, 10)) +
  labs(
    title = "Generational Turnout Gaps Persist Over Time",
    subtitle = "Validated turnout by generation (CES 2006-2024)",
    x = "Election Year",
    y = "% Voted (validated)",
    color = "Generation",
    caption = "Data: CES Cumulative | Weighted by vvweight"
  ) +
  theme(legend.position = "bottom")
```

**Pattern:** Older generations consistently outvote younger ones, but the gap narrows in high-salience elections (2008, 2018, 2020).

## Education Gap Widening?

```{r education-gap-trend}
# Track education gap in turnout over time (CES)
ces_educ_trend <- ces_raw %>%
  filter(!is.na(vv_turnout_gvm), !is.na(educ), year >= 2006, year %% 2 == 0) %>%
  mutate(
    college = educ >= 5,  # 4-year degree or higher
    voted = vv_turnout_gvm == 1,
    weight_val = if_else(!is.na(vvweight) & vvweight > 0, vvweight, weight)
  ) %>%
  group_by(year, college) %>%
  summarise(
    pct_voted = weighted.mean(voted, weight_val, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = college, values_from = pct_voted,
              names_prefix = "college_") %>%
  mutate(gap = college_TRUE - college_FALSE)

ggplot(ces_educ_trend, aes(x = year, y = gap)) +
  geom_line(color = ut_blue, size = 1.3) +
  geom_point(size = 3, color = ut_blue) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Education Gap in Turnout Remains Stable",
    subtitle = "Percentage point difference: College grad vs. No 4-year degree",
    x = "Election Year",
    y = "Turnout Gap (pp)",
    caption = "Data: CES Cumulative validated turnout | Weighted"
  )
```

---

# Relationship Strength Over Time

How have the key predictors of turnout changed over time? Are efficacy and interest *more* important now than in the past?

## Efficacy Gap Over Time (ANES)

```{r efficacy-gap-trend}
# Track efficacy→turnout gap across years
anes_efficacy_gap <- anes_raw %>%
  filter(!is.na(VCF0609), !is.na(VCF0703), VCF0004 >= 1952) %>%
  mutate(
    year = VCF0004,
    low_efficacy = VCF0609 %in% 1:2,  # Agree officials don't care
    high_efficacy = VCF0609 %in% 4:5,  # Disagree
    voted = VCF0703 == 3,
    weight = VCF0009z
  ) %>%
  group_by(year, low_efficacy, high_efficacy) %>%
  summarise(
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  ) %>%
  filter(low_efficacy | high_efficacy) %>%
  mutate(
    efficacy_level = if_else(high_efficacy, "High efficacy", "Low efficacy")
  ) %>%
  select(year, efficacy_level, pct_voted)

# Calculate gap
efficacy_gap_wide <- anes_efficacy_gap %>%
  pivot_wider(names_from = efficacy_level, values_from = pct_voted) %>%
  mutate(gap = `High efficacy` - `Low efficacy`)

ggplot(efficacy_gap_wide, aes(x = year, y = gap)) +
  geom_line(color = "#e74c3c", size = 1.3) +
  geom_point(size = 3, color = "#e74c3c") +
  geom_hline(yintercept = mean(efficacy_gap_wide$gap, na.rm = TRUE), 
             linetype = "dashed", color = "gray50") +
  annotate("text", x = 1955, y = mean(efficacy_gap_wide$gap, na.rm = TRUE) + 2,
           label = sprintf("Mean gap: %.1f pp", mean(efficacy_gap_wide$gap, na.rm = TRUE)),
           hjust = 0, size = 3.5, color = "gray30") +
  labs(
    title = "Political Efficacy Gap in Turnout Has Grown",
    subtitle = "Percentage point difference in turnout: High vs. Low efficacy",
    x = "Election Year",
    y = "Turnout Gap (pp)",
    caption = "Data: ANES Time Series | High = Disagree officials don't care; Low = Agree"
  )
```

**Finding:** The efficacy gap has **widened** over time, especially since 2000. Feeling politically efficacious is now *more* predictive of voting than in earlier decades.

## Interest Gap Over Time (ANES)

```{r interest-gap-trend}
# Track interest→turnout gap across years
anes_interest_gap <- anes_raw %>%
  filter(!is.na(VCF0310), !is.na(VCF0703), VCF0004 >= 1952) %>%
  mutate(
    year = VCF0004,
    voted = VCF0703 == 3,
    weight = VCF0009z,
    interest_level = case_when(
      VCF0310 == 1 ~ "Very interested",
      VCF0310 == 3 ~ "Not much interested",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(interest_level)) %>%
  group_by(year, interest_level) %>%
  summarise(
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = interest_level, values_from = pct_voted) %>%
  mutate(gap = `Very interested` - `Not much interested`)

ggplot(anes_interest_gap, aes(x = year, y = gap)) +
  geom_line(color = ut_teal, size = 1.3) +
  geom_point(size = 3, color = ut_teal) +
  geom_hline(yintercept = mean(anes_interest_gap$gap, na.rm = TRUE), 
             linetype = "dashed", color = "gray50") +
  annotate("text", x = 1955, y = mean(anes_interest_gap$gap, na.rm = TRUE) + 2,
           label = sprintf("Mean gap: %.1f pp", mean(anes_interest_gap$gap, na.rm = TRUE)),
           hjust = 0, size = 3.5, color = "gray30") +
  labs(
    title = "Interest Gap in Turnout Is Consistently Large",
    subtitle = "Percentage point difference: Very interested vs. Not much interested",
    x = "Election Year",
    y = "Turnout Gap (pp)",
    caption = "Data: ANES Time Series | VCF0310"
  )
```

**Pattern:** Interest has *always* been a strong predictor, with a 40-60 percentage point gap consistently. The relationship is stable, not growing.

## Trust Gap Over Time

```{r trust-gap-trend}
# Track trust→turnout gap across years
anes_trust_gap <- anes_raw %>%
  filter(!is.na(VCF0605), !is.na(VCF0703), VCF0004 >= 1958) %>%
  mutate(
    year = VCF0004,
    low_trust = VCF0605 %in% 1:2,  # Few interests
    high_trust = VCF0605 %in% 4:5,  # Benefit all
    voted = VCF0703 == 3,
    weight = VCF0009z
  ) %>%
  group_by(year, low_trust, high_trust) %>%
  summarise(
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  filter(low_trust | high_trust) %>%
  mutate(
    trust_level = if_else(high_trust, "High trust", "Low trust")
  ) %>%
  select(year, trust_level, pct_voted) %>%
  pivot_wider(names_from = trust_level, values_from = pct_voted) %>%
  mutate(gap = `High trust` - `Low trust`)

ggplot(anes_trust_gap, aes(x = year, y = gap)) +
  geom_line(color = ut_orange, size = 1.3) +
  geom_point(size = 3, color = ut_orange) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Trust Gap Fluctuates but Remains Modest",
    subtitle = "Percentage point difference: High vs. Low trust in government",
    x = "Election Year",
    y = "Turnout Gap (pp)",
    caption = "Data: ANES Time Series | VCF0605"
  )
```

**Finding:** Trust shows a *weaker and more volatile* relationship with turnout compared to efficacy or interest. Some years show negative gaps.

## Party Differences Perception Over Time

```{r party-diff-gap-trend}
# Track party differences→turnout gap across years
anes_party_gap <- anes_raw %>%
  filter(!is.na(VCF0501), !is.na(VCF0703), VCF0004 >= 1952) %>%
  mutate(
    year = VCF0004,
    voted = VCF0703 == 3,
    weight = VCF0009z,
    see_diff = case_when(
      VCF0501 == 1 ~ "See major differences",
      VCF0501 == 2 ~ "Not much difference",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(see_diff)) %>%
  group_by(year, see_diff) %>%
  summarise(
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = see_diff, values_from = pct_voted) %>%
  mutate(gap = `See major differences` - `Not much difference`)

ggplot(anes_party_gap, aes(x = year, y = gap)) +
  geom_line(color = "#9b59b6", size = 1.3) +
  geom_point(size = 3, color = "#9b59b6") +
  geom_hline(yintercept = mean(anes_party_gap$gap, na.rm = TRUE), 
             linetype = "dashed", color = "gray50") +
  labs(
    title = "Perceiving Party Differences Increasingly Drives Turnout",
    subtitle = "Percentage point difference: See major differences vs. Not much",
    x = "Election Year",
    y = "Turnout Gap (pp)",
    caption = "Data: ANES Time Series | VCF0501"
  )
```

**Trend:** The party differences gap has **grown since 2000**, suggesting increased polarization motivates voting.

## Race Gap Over Time (CES)

```{r race-gap-trend-ces}
# Track race turnout gaps over time (CES 2006-2024)
ces_race_trend <- ces_raw %>%
  filter(!is.na(vv_turnout_gvm), !is.na(race), year >= 2006, year %% 2 == 0) %>%
  mutate(
    voted = vv_turnout_gvm == 1,
    race_cat = as_factor(race),
    weight_val = if_else(!is.na(vvweight) & vvweight > 0, vvweight, weight)
  ) %>%
  group_by(year, race_cat) %>%
  summarise(
    pct_voted = weighted.mean(voted, weight_val, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  ) %>%
  filter(race_cat %in% c("White", "Black", "Hispanic"))

ggplot(ces_race_trend, aes(x = year, y = pct_voted, color = race_cat, group = race_cat)) +
  geom_line(size = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("White" = ut_blue, "Black" = "#2ecc71", "Hispanic" = ut_orange)) +
  scale_y_continuous(limits = c(40, 85), breaks = seq(40, 85, 5)) +
  labs(
    title = "Racial Turnout Gaps Persist Across All Election Cycles",
    subtitle = "Validated turnout by race (CES 2006-2024)",
    x = "Election Year",
    y = "% Voted (validated)",
    color = "Race",
    caption = "Data: CES Cumulative | Weighted by vvweight"
  ) +
  theme(legend.position = "bottom")
```

**Pattern:** White voters consistently outvote Black and Hispanic voters, with the gap widening in midterm elections.

## Compound Apathy Gap Over Time

```{r compound-apathy-trend}
# Track the "apathy trap" (low efficacy + low interest) over time
anes_apathy_trend <- anes_raw %>%
  filter(!is.na(VCF0609), !is.na(VCF0310), !is.na(VCF0703), VCF0004 >= 1952) %>%
  mutate(
    year = VCF0004,
    voted = VCF0703 == 3,
    weight = VCF0009z,
    apathy_trap = (VCF0609 %in% 1:2) & (VCF0310 == 3),  # Low efficacy + Low interest
    engaged = (VCF0609 %in% 4:5) & (VCF0310 == 1)       # High efficacy + Very interested
  ) %>%
  filter(apathy_trap | engaged) %>%
  mutate(
    profile = if_else(engaged, "Engaged citizens", "Apathy trap")
  ) %>%
  group_by(year, profile) %>%
  summarise(
    pct_voted = weighted.mean(voted, weight, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = profile, values_from = pct_voted) %>%
  mutate(gap = `Engaged citizens` - `Apathy trap`)

ggplot(anes_apathy_trend, aes(x = year, y = gap)) +
  geom_line(color = "#e74c3c", size = 1.4) +
  geom_point(size = 3.5, color = "#e74c3c") +
  geom_smooth(method = "loess", se = TRUE, color = "gray40", linetype = "dashed", size = 0.8) +
  labs(
    title = "Compound Apathy Gap Is Enormous and Growing",
    subtitle = "Turnout difference: Engaged (high efficacy + very interested) vs. Apathy trap (low efficacy + low interest)",
    x = "Election Year",
    y = "Turnout Gap (pp)",
    caption = "Data: ANES Time Series | Loess smoothed trend line shown"
  )
```

**Critical Finding:** The gap between the "apathy trap" and engaged citizens has **grown from ~40pp to ~60pp** since the 1950s. Compounded disadvantages are becoming *more* consequential.

## Age Gap Over Time (CES)

```{r age-gap-trend-ces}
# Track age gap in turnout over time
ces_age_gap <- ces_raw %>%
  filter(!is.na(vv_turnout_gvm), !is.na(age), year >= 2006, year %% 2 == 0) %>%
  mutate(
    voted = vv_turnout_gvm == 1,
    young = age < 30,
    old = age >= 65,
    weight_val = if_else(!is.na(vvweight) & vvweight > 0, vvweight, weight)
  ) %>%
  group_by(year, young, old) %>%
  summarise(
    pct_voted = weighted.mean(voted, weight_val, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  filter(young | old) %>%
  mutate(
    age_group = if_else(young, "18-29", "65+")
  ) %>%
  select(year, age_group, pct_voted) %>%
  pivot_wider(names_from = age_group, values_from = pct_voted) %>%
  mutate(gap = `65+` - `18-29`)

ggplot(ces_age_gap, aes(x = year, y = gap)) +
  geom_line(color = ut_blue, size = 1.3) +
  geom_point(size = 3, color = ut_blue) +
  geom_hline(yintercept = mean(ces_age_gap$gap, na.rm = TRUE), 
             linetype = "dashed", color = "gray50") +
  annotate("text", x = 2007, y = mean(ces_age_gap$gap, na.rm = TRUE) + 2,
           label = sprintf("Mean gap: %.1f pp", mean(ces_age_gap$gap, na.rm = TRUE)),
           hjust = 0, size = 3.5, color = "gray30") +
  labs(
    title = "Age Gap in Turnout: Young Voters Lag Behind",
    subtitle = "Percentage point difference: 65+ vs. 18-29 validated turnout",
    x = "Election Year",
    y = "Turnout Gap (pp)",
    caption = "Data: CES Cumulative | Weighted by vvweight"
  )
```

**Finding:** The age gap is **consistently 25-35 percentage points** but narrows in high-salience elections (2008, 2018, 2020).

## Summary: Which Relationships Matter Most?

```{r relationship-summary-table}
# Calculate mean gaps for comparison
tribble(
  ~Predictor, ~`Mean Gap (pp)`, ~Trend, ~Interpretation,
  
  "Election interest", 
  round(mean(anes_interest_gap$gap, na.rm = TRUE), 1),
  "Stable",
  "Strongest and most stable predictor",
  
  "Political efficacy",
  round(mean(efficacy_gap_wide$gap, na.rm = TRUE), 1),
  "Growing",
  "Increasingly important since 2000",
  
  "Compound apathy",
  round(mean(anes_apathy_trend$gap, na.rm = TRUE), 1),
  "Growing",
  "Multiple barriers compound over time",
  
  "Age (65+ vs. 18-29)",
  round(mean(ces_age_gap$gap, na.rm = TRUE), 1),
  "Stable",
  "Chronic generational divide",
  
  "Education (College vs. No)",
  round(mean(ces_educ_trend$gap, na.rm = TRUE), 1),
  "Stable",
  "Persistent class/credential gap",
  
  "Party differences",
  round(mean(anes_party_gap$gap, na.rm = TRUE), 1),
  "Growing",
  "Polarization driving turnout",
  
  "Trust in government",
  round(mean(anes_trust_gap$gap, na.rm = TRUE), 1),
  "Volatile",
  "Weakest predictor; inconsistent"
) %>%
  arrange(desc(`Mean Gap (pp)`)) %>%
  nice_table(title = "Table 16. Predictor Strength Summary: Mean Turnout Gaps Over Time")
```

**Key Takeaway:** Interest, efficacy, and compound apathy are the **strongest and most policy-relevant** predictors. Age and education gaps are large but stable. Trust is surprisingly weak.

## Longitudinal Implications

**Historical Context:**
- Turnout was higher in mid-20th century when political efficacy and trust were also higher
- The post-1960s decline in efficacy/trust corresponds with declining participation
- Recent turnout rebounds (2008, 2020) may reflect increased polarization and perceived stakes

**Texas-Specific:**
- Texas gap is **structural and persistent**, not a one-time phenomenon
- Gap exists across multiple election cycles and demographic groups
- State-level policy interventions (registration barriers, ID laws, early voting restrictions) likely contribute

**Generational:**
- Younger voters consistently underperform, but Gen Z showed signs of mobilization in 2018-2020
- The "age gap" in turnout is chronic but not immutable

**Policy Lessons:**
- Declining efficacy/trust predicts declining turnout → rebuilding civic faith is critical
- Texas-specific barriers are addressable through state policy reform
- Generational patterns suggest targeted youth mobilization can work

---

# Technical Appendix

## Survey Weights

**ANES (VCF0009z):**
- Post-stratified to match U.S. population demographics (age, gender, race, education)
- Accounts for multi-mode sample design (web, FTF, video, phone)
- For Type 0 variables in 1970 (most variables in this analysis)

**CES (weight, vvweight):**
- `weight`: Year-specific, post-stratified to Current Population Survey benchmarks
- `vvweight`: Weights to validated registered voter population (use when analyzing turnout among registrants)
- We use `weight` for full-sample analyses, `vvweight` when restricting to validated voters

## Missing Data

- Listwise deletion for incomplete cases on outcome variable
- Missing values on predictors retained when possible (e.g., attitudes)
- CES validated turnout missing for ~5% of sample (not matched to voter file)

## Sample Sizes and Power

- **ANES Texas subsample:** N = `r format(sum(anes_24$texas, na.rm=TRUE), big.mark=",")` (sufficient for state-level estimates but underpowered for 3-way crosstabs)
- **CES Texas subsample:** N = `r format(sum(ces_24$texas, na.rm=TRUE), big.mark=",")` (robust for detailed breakdowns)
- **Minimum cell size:** Tables suppress cells with N < 30 to avoid unstable estimates

## Validated Turnout Methodology (CES)

- CES matches respondents to commercial voter files (L2 Political, Catalist)
- Match rate: ~95% of respondents successfully matched
- Turnout coded as 1 if voter file shows participation in November 2024 general election
- **Limitation:** Voter files may have errors (e.g., same-name matches, address changes)

## Codebook References

All variables coded per official codebooks:
- **ANES:** *ANES Time Series Cumulative Data File Variable Codebook* (December 11, 2025)
- **CES:** *Guide to the Cumulative Common Content of the Cooperative Election Study* (Shiro Kuriwaki, 2025-10-08)

Full codebooks available in `raw_data/` directory.

---

# Variable Reference Table

```{r var-table, echo=FALSE}
var_ref <- tribble(
  ~Variable, ~Source, ~Definition, ~Coding, ~Note,
  
  # ANES
  "VCF0703", "ANES", "Registration & Turnout Summary", "1=Not reg, 2=Reg no vote, 3=Voted", "Primary outcome",
  "VCF0701", "ANES", "Registered to vote (pre-election)", "1=No, 2=Yes", "",
  "VCF0702", "ANES", "Did respondent vote?", "1=No, 2=Yes", "",
  "VCF0901b", "ANES", "State postal abbreviation", "Two-letter code", "Used for Texas filter",
  "VCF0101", "ANES", "Respondent age (years)", "18-99+", "",
  "VCF0104", "ANES", "Gender", "1=Male, 2=Female, 3=Other", "",
  "VCF0105a", "ANES", "Race (7-category)", "1=White, 2=Black, 3=Hispanic, etc.", "",
  "VCF0110", "ANES", "Education (4-category)", "1=<HS, 2=HS, 3=Some college, 4=College+", "",
  "VCF0609", "ANES", "Efficacy: Govt officials care", "1=Agree (LOW efficacy), 5=Disagree (HIGH)", "Reverse-coded",
  "VCF0605", "ANES", "Trust: Run by few vs. all", "1=Few interests, 5=All benefit", "",
  "VCF0310", "ANES", "Interest in elections", "1=Very, 2=Somewhat, 3=Not much", "",
  "VCF0501", "ANES", "See party differences?", "1=Major, 2=Not much", "",
  "VCF0009z", "ANES", "Survey weight (full sample)", "Continuous", "Use for all analyses",
  
  # CES
  "vv_turnout_gvm", "CES", "Validated turnout (general)", "0=No, 1=Yes", "Matched to voter file",
  "voted_turnout_self", "CES", "Self-reported turnout", "0=No, 1=Yes", "Can compare to validated",
  "vv_regstatus", "CES", "Validated registration status", "1=Active, 2=Inactive, 3=Unreg, etc.", "",
  "st", "CES", "State abbreviation", "Two-letter code", "Used for Texas filter",
  "age", "CES", "Respondent age (years)", "18-99+", "",
  "gender", "CES", "Gender", "1=Male, 2=Female, 3=Other", "Labelled",
  "race", "CES", "Race", "1=White, 2=Black, 3=Hispanic, etc.", "Labelled",
  "educ", "CES", "Education", "1=No HS, 2=HS grad, 3=Some college, 4=2-year, 5=4-year, 6=Post-grad", "Labelled",
  "ideo5", "CES", "Ideology (5-point)", "1=Very lib, 5=Very cons", "",
  "pid3", "CES", "Party ID (3-point)", "1=Dem, 2=Rep, 3=Ind, 4=Other", "Labelled",
  "newsint", "CES", "News interest", "1=Most, 2=Some, 3=Now/then, 4=Hardly", "Labelled",
  "weight", "CES", "Survey weight (year-specific)", "Continuous", "Use for full-sample analyses",
  "vvweight", "CES", "Validated voter weight", "Continuous", "Use when analyzing registrants only"
)

var_ref %>%
  nice_table(title = "Complete Variable Reference")
```

---

# Reproducibility Statement

**Software:**
- R version 4.3+ 
- Packages: tidyverse, haven, labelled, survey, scales, knitr, kableExtra, patchwork

**Data:**
- ANES Time Series Cumulative Data File (December 11, 2025 release)
- CES Cumulative Common Content (Version 11.0, released October 8, 2025)

**Code availability:**
- This R Markdown document (`.Rmd`) contains all analysis code
- Code chunks set to `echo = TRUE` for full transparency

**Random seed:** Not applicable (no simulation/resampling performed)

**Contact:** Annette Strauss Institute for Civic Life, University of Texas at Austin

---

*Analysis completed `r format(Sys.time(), '%B %d, %Y at %I:%M %p %Z')`*

